---
title: "STAD and COAD analysis"
author: "nvolkova"
date: "Apr 2017"
output:
  pdf_document: default
  html_notebook: default
  html_document: default
  word_document: default
---

```{r setup, echo=F, eval=F}
# Ensure your working directory is MMR
knitr::opts_knit$set(root.dir = '/Users/nvolkova/MMR/')
```

This is analysis of mutational signatures in COAD and STAD datasets from [ICGC](http://icgc.org) for MMR paper ([Meier, Volkova et al. 2017](https://www.biorxiv.org/content/biorxiv/early/2017/06/13/149153.full.pdf)).

## Get the data and signatures and contributions

Preparation

```{r, message=F, warning=F, eval=F}
source('plotting functions/plot_sigs.R')
source("nmSolve.R")
source("plotting functions/plot_decomposition.R")
source('plotting functions/scatterpie.R')
library(tsne)
library(NMF)
library(devtools)
#devtools::install_github("mg14/mg14") # for plotting purposes
library(mg14)
library(ggplot2)
library(reshape2)
library(VariantAnnotation)
library(deconstructSigs)
library(Biostrings)
library(GenomicRanges)
library(dplyr)
library(gam)
```

### Functions for plotting:

`plot_decomposition` - function for plotting decompositions of mutational profiles over different signatures
Requires: `ggplot2`, `reshape2` libraries.
Arguments:
- `decomposition` - matrix of fractions, samples x signatures
- `mm` - matrix of mutational counts of different types, samples x 96 (or 104) mutation types
- `intnames` - character vector of names of samples to plot (should be present in decomposition and mm's rownames)
- `col` - vector of colours to use for signatures
- `circle` - TRUE or FALSE, make it a circular barplot ('nautilus-plot') or a normal one
- `size` - regulates the size of x-axis text (sample names)
- `axis.size` - regulates the size og legend text and y-axis text (signature names and numbers on y scale)
```{r,eval=F}
plot_decomposition <- function(decomposition, mm, intnames, col,circle=F,size=6,axis.size=10) {
  for (i in 1:nrow(decomposition))
    decomposition[i,] = decomposition[i,] / sum(decomposition[i,])
  new.cont.mat <- t(decomposition[intnames,])
  for (y in colnames(new.cont.mat)) {
    new.cont.mat[,y] = new.cont.mat[,y] * rowSums(mm)[y]
  }
  m.new.cont.mat <- melt(new.cont.mat)
  colnames(m.new.cont.mat) = c("Signature","Sample","Contribution")
  order(rowSums(mm),decreasing = F) -> sampleorder
  names(sampleorder) <- row.names(mm)[sampleorder]
  plot = ggplot(m.new.cont.mat, aes(x = factor(Sample,levels=names(sampleorder)), 
                                    y = Contribution, fill = factor(Signature,levels=colnames(decomposition)), order = Sample)) + 
    geom_bar(stat = "identity", colour = "black") + 
    labs(x = "", y = "Absolute contribution \n (no. mutations)") + 
    theme_bw() + 
    scale_fill_manual(name="",values=col) + 
    theme(panel.grid.minor.x = element_blank(), panel.grid.major.x = element_blank()) + 
    theme(panel.grid.minor.y = element_blank(), panel.grid.major.y = element_blank()) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1,size=size),
          axis.text.y = element_text(size=axis.size),
          legend.text=element_text(size=axis.size))
  if (circle) {
    new.cont.mat <- t(decomposition[intnames,])
    for (y in colnames(new.cont.mat)) {
      new.cont.mat[,y] = new.cont.mat[,y] * log10(rowSums(mm)[y])
    }
    m.new.cont.mat <- melt(new.cont.mat)
    colnames(m.new.cont.mat) = c("Signature","Sample","Contribution")
    plot = ggplot(m.new.cont.mat, aes(x = factor(Sample,levels=names(sampleorder)), 
                                                y = Contribution, fill = factor(Signature,levels=colnames(decomposition)), order = Sample)) + 
    geom_bar(stat = "identity", colour = "black") + 
    theme_bw() + 
    scale_fill_manual(name="",values=col) + 
    theme(panel.grid.minor.x = element_blank(), panel.grid.major.x = element_blank()) + 
    theme(panel.grid.minor.y = element_blank(), panel.grid.major.y = element_blank()) +
    theme(axis.text.x = element_text(size=size),
          axis.text.y = element_text(size=axis.size),
          legend.text=element_text(size=axis.size)) + 
    coord_polar() + 
    labs(x = "", y = "Absolute contribution \n (no. mutations, log10)")
  }
  plot
}
```

`plot_sig_wb` - function for plotting 96-long signatures of relative contributions of different mutation types: different design
Requires: `ggplot2` and `reshape2` libraries.
Arguments:
* `mut_matrix` - matrix of signatures (or profiles), samples x signatures
* `colors` - colors to use for 6 mutation classes
* `ymin`, `ymax` - max and min value on the plot
* `flip` - FALSE or TRUE, flips the plot vertically
* `size` - size of mutation class titles
```{r,eval=F}
plot_sig_wb <- function (mut_matrix, colors, ymin=0,ymax = 0.15,flip=F,size=12) # plotting 96-signatures with no lines
{
  C_TRIPLETS = c(
    "ACA", "ACC", "ACG", "ACT",
    "CCA", "CCC", "CCG", "CCT",
    "GCA", "GCC", "GCG", "GCT",
    "TCA", "TCC", "TCG", "TCT")
  
  T_TRIPLETS = c(
    "ATA", "ATC", "ATG", "ATT",
    "CTA", "CTC", "CTG", "CTT",
    "GTA", "GTC", "GTG", "GTT",
    "TTA", "TTC", "TTG", "TTT")
  
  IND_TRIPLETS = c(
    "A*A", "A*C", "A*G", "A*T",
    "C*A", "C*C", "C*G", "C*T",
    "G*A", "G*C", "G*G", "G*T",
    "T*A", "T*C", "T*G", "T*T")
  
  TRIPLETS_96 = c(rep(C_TRIPLETS, 3), rep(T_TRIPLETS, 3))
  TRIPLETS_112 = c(rep(C_TRIPLETS, 3), rep(T_TRIPLETS, 3),IND_TRIPLETS)
  norm_mut_matrix = apply(mut_matrix, 2, function(x) x/sum(x))
  if (missing(colors)) {
    colors = c("#2EBAED", "#000000", "#DE1C14",
               "#D4D2D2", "#ADCC54", "#F0D0CE")
  }
  context = TRIPLETS_96
  substitution = rep(c("C>A","C>G","C>T","T>A","T>C","T>G"), each = 16)
  substring(context, 2, 2) = "*"
  df = data.frame(substitution = substitution, context = context)
  rownames(norm_mut_matrix) = NULL
  if (flip==T) norm_mut_matrix = -norm_mut_matrix
  df2 = cbind(df, as.data.frame(norm_mut_matrix))
  df3 = melt(df2, id.vars = c("substitution", "context"))
  value = NULL
  if (ymax>0) breaks = c(0,0.1) else breaks = c(-0.1,0)
  if (ymax>0) lbls = c('0','0.1') else lbls = c('0.1','0')
  plot = ggplot(data = df3, aes(x = context, y = value, fill = substitution, width = 0.6)) +
    geom_bar(stat = "identity", colour = "black",size = 0.2) +
    scale_fill_manual(values = colors) + facet_grid(variable ~ substitution) +
    ylab("Relative contribution") + coord_cartesian(ylim = c(ymin,ymax)) + 
    guides(fill = FALSE) + theme_bw() + scale_y_continuous(breaks = breaks, labels = lbls) +
    theme(text = element_text(family='ArialMT'),
          axis.title.y = element_text(size = 14,vjust = 1), 
          axis.text.y = element_text(size = 8), 
          axis.title.x = element_text(size = 14), 
          axis.text.x = element_text(size = 8, angle = 90, vjust = 0.4), 
          strip.text.x = element_text(size = 16), 
          strip.text.y = element_text(size = size), 
          panel.grid.major.x = element_blank(), 
          strip.background = element_blank(), 
          panel.border = element_rect(colour="white"),
          panel.spacing = unit(0.1,'lines'))
  return(plot)
}
```

`plot_profiles` - function for plotting 96-long mutational profiles (in absolute numbers).
Requires: `ggplot2` and `reshape2` libraries.
Arguments:
- `mut_matrix` - matrix of signatures (or profiles), samples x signatures
- `colors` - colors to use for 6 mutation classes
```{r,eval=F}
plot_profiles <- function (mut_matrix, colors) # plot mutational profiles without normalizing
{
  C_TRIPLETS = c(
    "ACA", "ACC", "ACG", "ACT",
    "CCA", "CCC", "CCG", "CCT",
    "GCA", "GCC", "GCG", "GCT",
    "TCA", "TCC", "TCG", "TCT")
  
  T_TRIPLETS = c(
    "ATA", "ATC", "ATG", "ATT",
    "CTA", "CTC", "CTG", "CTT",
    "GTA", "GTC", "GTG", "GTT",
    "TTA", "TTC", "TTG", "TTT")
  
  TRIPLETS_96 = c(rep(C_TRIPLETS, 3), rep(T_TRIPLETS, 3))
  norm_mut_matrix = apply(mut_matrix, 2, function(x) x/sum(x))
  if (missing(colors)) {
    colors = c(
      "#2EBAED", "#000000", "#DE1C14",
      "#D4D2D2", "#ADCC54", "#F0D0CE")
  }
  if (length(colors) != 6) {
    stop("Provide colors vector with length 6")
  }
  context = TRIPLETS_96
  substitution = rep(c("C>A","C>G","C>T","T>A","T>C","T>G"), each = 16)
  substring(context, 2, 2) = "*"
  df = data.frame(substitution = substitution, context = context)
  df2 = cbind(df, as.data.frame(mut_matrix))
  df3 = melt(df2, id.vars = c("substitution", "context"))
  value = NULL
  plot = ggplot(data = df3, aes(x = context, y = value, fill = substitution, width = 0.6)) +
    geom_bar(stat = "identity", colour = "black",size = 0.2) +
    scale_fill_manual(values = colors) + facet_grid(variable ~ substitution) +
    ylab("Relative contribution") + coord_cartesian(ylim = c(0,max(mut_matrix))) + 
    guides(fill = FALSE) + theme_bw() +
    theme(axis.title.y = element_text(size = 14,vjust = 1), axis.text.y = element_text(size = 10), axis.title.x = element_text(size = 14), 
          axis.text.x = element_text(size = 7, angle = 90, vjust = 0.4), 
          strip.text.x = element_text(size = 11), strip.text.y = element_text(size = 13), 
          panel.grid.major.x = element_blank())
  return(plot)
}
```

`plot_sig_104` - function for plotting 104-long signatures of relative contributions of different mutation types.
Requires: `ggplot2` and `reshape2` libraries.
Arguments:
* `mut_matrix` - matrix of signatures (or profiles), mutation types x signatures
* `colors` - colors to use for 8 mutation classes
* `size` - size of mutation class titles
* `ymax` - maximal value on the plot
```{r,eval=F}
plot_sig_104 <- function(mut_matrix,colors=NA,size=8,ymax=0.2) { # plot 104-signatures
  
  C_TRIPLETS = c(
    "ACA", "ACC", "ACG", "ACT",
    "CCA", "CCC", "CCG", "CCT",
    "GCA", "GCC", "GCG", "GCT",
    "TCA", "TCC", "TCG", "TCT")
  
  T_TRIPLETS = c(
    "ATA", "ATC", "ATG", "ATT",
    "CTA", "CTC", "CTG", "CTT",
    "GTA", "GTC", "GTG", "GTT",
    "TTA", "TTC", "TTG", "TTT")
  
  if (is.na(colors)) colors=c("#2EBAED", "#000000", "#DE1C14","orange","purple","#D4D2D2", "#ADCC54", "#F0D0CE")
  
  types <- c(rep(C_TRIPLETS,3), rep(T_TRIPLETS,3))
  norm_mut_matrix = apply(mut_matrix, 2, function(x) x/sum(x))
  df = as.data.frame(norm_mut_matrix)
  df$Type = c(rep(c("C>A","C>G","C>T","T>A","T>C","T>G"),each=16),rep("INS",4),rep("DEL",4))
  df$Base = c(types,rep(c("A","C","G","T"),2))
  substring(df$Base[nchar(df$Base)>1], 2, 2) = "*"
  #colnames(df)[1:ncol(mut_matrix)] = paste0("Signature.",c(1:ncol(mut_matrix)),sep="")
  df2 = melt(df,id.vars = c("Type","Base"))
  df2$Type_f = factor(df2$Type, levels=c("C>A","C>G","C>T","T>A","T>C","T>G","DEL","INS"))
  plot = ggplot(data = df2, aes(x = Base, y = value, fill = Type, width = 0.6)) +
    geom_bar(stat = "identity", colour = "black",size = 0.2) +
    scale_fill_manual(values = colors) + 
    facet_grid(variable ~ Type_f,scales = "free_x") +
    ylab("Relative contribution") + coord_cartesian(ylim = c(0,ymax)) + 
    guides(fill = FALSE) + theme_bw() + scale_y_continuous(breaks = c(0,0.15)) +
    theme(text = element_text(family='ArialMT'),
          axis.title.y = element_text(size = 14,vjust = 1), 
          axis.text.y = element_text(size = 8), 
          axis.title.x = element_text(size = 14), 
          axis.text.x = element_text(size = 8, angle = 90, vjust = 0.4), 
          strip.text.x = element_text(size = 16), 
          strip.text.y = element_text(size = size), 
          panel.grid.major.x = element_blank(), 
          strip.background = element_blank(), 
          panel.border = element_rect(colour="white"),
          panel.spacing = unit(0.1,'lines'))
  return(plot)
}
```

`plot_sig_104_CI` - function for plotting 104-long signatures of relative contributions of different mutation types with confidence intervals.
Requires: `ggplot2` and `reshape2` libraries.
Arguments:
* `mut_matrix` - matrix of signatures (or profiles), mutation types x signatures
* `mut_matrix_lower` - matrix of lower CI values for signatures, mutation types x signatures
* `mut_matrix_upper` - matrix of upper CI values for signatures, mutation types x signatures
* `colors` - colors to use for 8 mutation classes
* `size` - size of mutation class titles
* `ymax` - maximal value on the plot
```{r,eval=F}
plot_sig_104_CI <- function(mut_matrix,mut_matrix_lower,mut_matrix_upper,size=8,ymax=0.2,colors) { # plot 104-signatures with confidence intervals
  
  colors = c("#2EBAED", "#000000", "#DE1C14","orange","purple","#D4D2D2", "#ADCC54", "#F0D0CE")
  
  
  C_TRIPLETS = c(
    "ACA", "ACC", "ACG", "ACT",
    "CCA", "CCC", "CCG", "CCT",
    "GCA", "GCC", "GCG", "GCT",
    "TCA", "TCC", "TCG", "TCT")
  
  T_TRIPLETS = c(
    "ATA", "ATC", "ATG", "ATT",
    "CTA", "CTC", "CTG", "CTT",
    "GTA", "GTC", "GTG", "GTT",
    "TTA", "TTC", "TTG", "TTT")
  
  types <- c(rep(C_TRIPLETS,3), rep(T_TRIPLETS,3))
  df = as.data.frame(mut_matrix)
  df$Type = c(rep(c("C>A","C>G","C>T","T>A","T>C","T>G"),each=16),rep("INS",4),rep("DEL",4))
  df$Base = c(types,rep(c("A","C","G","T"),2))
  substring(df$Base[nchar(df$Base)>1], 2, 2) = "*"
  df2 = melt(df,id.vars = c("Type","Base"))
  df2$Type_f = factor(df2$Type, levels=c("C>A","C>G","C>T","T>A","T>C","T>G","DEL","INS"))
  plot = ggplot(data = df2, aes(x = Base, y = value, fill = Type, width = 0.6)) +
    geom_bar(stat = "identity", colour = "black",size = 0.2) +
    scale_fill_manual(values = colors) + 
    facet_grid(variable ~ Type_f,scales = "free_x") +
    ylab("Relative contribution") + coord_cartesian(ylim = c(0,ymax)) + 
    guides(fill = FALSE) + theme_bw() +
    theme(axis.title.y = element_text(size = 14,vjust = 1), axis.text.y = element_text(size = 6), axis.title.x = element_text(size = 14), 
          axis.text.x = element_text(size = 6, angle = 90, vjust = 0.4), 
          strip.text.x = element_text(size = 10), strip.text.y = element_text(size = size), 
          panel.grid.major.x = element_blank())
  
  mut_matrix_lower <- mut_matrix_lower[row.names(mut_matrix),]
  mut_matrix_upper <- mut_matrix_upper[row.names(mut_matrix),]
  rownames(mut_matrix_lower) = rownames(mut_matrix_upper) = NULL
  df_lower = as.data.frame(mut_matrix_lower)
  df_upper = as.data.frame(mut_matrix_upper)
  df_lower$Type = c(rep(c("C>A","C>G","C>T","T>A","T>C","T>G"),each=16),rep("INS",4),rep("DEL",4))
  df_lower$Base = c(types,rep(c("A","C","G","T"),2))
  df_upper$Type = c(rep(c("C>A","C>G","C>T","T>A","T>C","T>G"),each=16),rep("INS",4),rep("DEL",4))
  df_upper$Base = c(types,rep(c("A","C","G","T"),2))
  df_lower = melt(df_lower, id.vars = c("Type","Base"))
  df_upper = melt(df_upper, id.vars = c("Type","Base"))
  df3 <- cbind(df2, value_min = df_lower$value, value_max = df_upper$value)
  df3$Type_f = factor(df3$Type, levels=c("C>A","C>G","C>T","T>A","T>C","T>G","DEL","INS"))
  plot2 = plot + geom_pointrange(data=df3, aes(ymin=value,ymax=value_max,colour=factor(Type)), fatten = 0.01, size=0.5, show.legend = F, colour="white") +
    scale_color_manual(values=colors) +
    geom_pointrange(data=df3, aes(ymin=value_min,ymax=value), size=0.5, fatten = 0.01,show.legend = F, colour="white")
  
  return(plot2)
}
```

`plot_profiles_104` - function for plotting 104-long mutational profiles (normalized, i.e. in relative contribution).
Requires: `ggplot2` and `reshape2` libraries.
Arguments:
* `mut_matrix` - matrix of profiles, samples x mutation counts
* `colors` - colors to use for 8 mutation classes
* `boxplot` - FALSE or TRUE, merges all the values across profiles into a boxplot
* `normalize` - FALSE or TRUE, turns absolute numbers into relative contributions
* `size` - regulates font size of the text on the plot
```{r,eval=F}
plot_profiles_104 <- function (mut_matrix, colors=NA, boxplot=F, normalize=F, size=6) # plot 104-long mutational profiles
{
  
  if (is.na(colors)) colors = c("#2EBAED", "#000000", "#DE1C14","orange","purple","#D4D2D2", "#ADCC54", "#F0D0CE")
  
  C_TRIPLETS = c(
    "ACA", "ACC", "ACG", "ACT",
    "CCA", "CCC", "CCG", "CCT",
    "GCA", "GCC", "GCG", "GCT",
    "TCA", "TCC", "TCG", "TCT")
  
  T_TRIPLETS = c(
    "ATA", "ATC", "ATG", "ATT",
    "CTA", "CTC", "CTG", "CTT",
    "GTA", "GTC", "GTG", "GTT",
    "TTA", "TTC", "TTG", "TTT")
  
  types <- c(rep(C_TRIPLETS,3), rep(T_TRIPLETS,3))
  if (normalize) {
    mut_matrix <- apply(mut_matrix, 2, function(x) x/sum(x))
  }
  df = as.data.frame(mut_matrix)
  df$Type = c(rep(c("C>A","C>G","C>T","T>A","T>C","T>G"),each=16),rep("INS",4),rep("DEL",4))
  df$Base = c(types,rep(c("A","C","G","T"),2))
  substring(df$Base[nchar(df$Base)>1], 2, 2) = "*"
  df2 = melt(df,id.vars = c("Type","Base"))
  df2$Type_f = factor(df2$Type, levels=c("C>A","C>G","C>T","T>A","T>C","T>G","DEL","INS"))
  if (boxplot) {
    plot = ggplot(data = df2, aes(x = Base, y = value, fill=Type)) +
      geom_boxplot() +
      scale_fill_manual(values = colors) + 
      facet_grid(. ~ Type_f, scales="free") +
      ylab("Mutation counts") + 
      guides(fill = FALSE) + theme_bw() +
      theme(axis.title.y = element_text(size = 20,vjust = 1), axis.text.y = element_text(size = 16), axis.title.x = element_text(size = 20), 
            axis.text.x = element_text(size = 8, angle = 90, vjust = 0.4), 
            strip.text.x = element_text(size = size), strip.text.y = element_text(size = size), 
            panel.grid.major.x = element_blank())
  } else {
    plot = ggplot(data = df2, aes(x = Base, y = value, fill = Type, width = 0.6)) +
      geom_bar(stat = "identity", colour = "black",size = 0.2) +
      scale_fill_manual(values = c("#2EBAED", "#000000", "#DE1C14","orange","purple","#D4D2D2", "#ADCC54", "#F0D0CE")) + 
      facet_grid(variable ~ Type_f,scales = "free_x") +
      ylab("Mutation counts") + coord_cartesian(ylim = c(0,max(mut_matrix))) + 
      guides(fill = FALSE) + theme_bw() +
      theme(axis.title.y = element_text(size = 14,vjust = 1), axis.text.y = element_text(size = 6), axis.title.x = element_text(size = 14), 
            axis.text.x = element_text(size = 6, angle = 90, vjust = 0.4), 
            strip.text.x = element_text(size = 6), strip.text.y = element_text(size = 6), 
            panel.grid.major.x = element_blank())
  }
  return(plot)
}
```

`plot_104_profile_CI` - function for plotting 104-long mutational profiles (in absolute numbers) with confidence intervals.
Requires: `ggplot2` and `reshape2` libraries.
Arguments:
* `mut_matrix` - matrix of profiles, samples x mutation counts
* `colors` - colors to use for 8 mutation classes
```{r,eval=F}
plot_104_profile_CI <- function (mut_matrix, mut_matrix_lower = NULL, 
                                mut_matrix_upper = NULL,  
                                colors=c("deepskyblue2","black","red3","orange","purple","grey","olivedrab3","pink"),size=6,ymax=0.25) # plot 104-long profiles with confidence intervals
{
  C_TRIPLETS = c(
    "ACA", "ACC", "ACG", "ACT",
    "CCA", "CCC", "CCG", "CCT",
    "GCA", "GCC", "GCG", "GCT",
    "TCA", "TCC", "TCG", "TCT")
  
  T_TRIPLETS = c(
    "ATA", "ATC", "ATG", "ATT",
    "CTA", "CTC", "CTG", "CTT",
    "GTA", "GTC", "GTG", "GTT",
    "TTA", "TTC", "TTG", "TTT")
  
  
  norm_mut_matrix = mut_matrix
  mult <- colSums(norm_mut_matrix)
  norm_mut_matrix = apply(mut_matrix, 2, function(x) x/sum(x))
  types <- c(rep(C_TRIPLETS,3), rep(T_TRIPLETS,3))
  df = as.data.frame(norm_mut_matrix)
  df$Type = c(rep(c("C>A","C>G","C>T","T>A","T>C","T>G"),each=16),rep("INS",4),rep("DEL",4))
  df$Base = c(types,rep(c("A","C","G","T"),2))
  df2 = melt(df,id.vars = c("Type","Base"))
  df2$Type_f = factor(df2$Type, levels=c("C>A","C>G","C>T","T>A","T>C","T>G","DEL","INS"))
  plot = ggplot(data = df2, aes(x = Base, y = value, fill = Type, width = 0.6)) +
    geom_bar(stat = "identity", colour = "black",size = 0.2) +
    scale_fill_manual(values = colors) + 
    facet_grid(variable ~ Type_f,scales = "free_x") +
    ylab("Mutation counts") + coord_cartesian(ylim = c(0,ymax)) + 
    guides(fill = FALSE) + theme_bw() +
    theme(axis.title.y = element_text(size = 16,vjust = 1), axis.text.y = element_text(size = size), axis.title.x = element_text(size = 16), 
          axis.text.x = element_text(size = size, angle = 90, vjust = 0.4), 
          strip.text.x = element_text(size = size), strip.text.y = element_text(size = size), 
          panel.grid.major.x = element_blank())
  
    mut_matrix_lower <- mut_matrix_lower[row.names(mut_matrix),]
    mut_matrix_upper <- mut_matrix_upper[row.names(mut_matrix),]
    norm_mut_matrix_lower <- mut_matrix_lower
    norm_mut_matrix_upper <- mut_matrix_upper
    for (i in 1:ncol(norm_mut_matrix_lower)) {
      norm_mut_matrix_lower[,i] = norm_mut_matrix_lower[,i] / mult[colnames(norm_mut_matrix_lower)[i]]
      norm_mut_matrix_upper[,i] = norm_mut_matrix_upper[,i] / mult[colnames(norm_mut_matrix_lower)[i]]
    }
    rownames(norm_mut_matrix_upper) = rownames(norm_mut_matrix_lower) = NULL
    df_lower = as.data.frame(norm_mut_matrix_lower)
    df_upper = as.data.frame(norm_mut_matrix_upper)
    df_lower$Type = c(rep(c("C>A","C>G","C>T","T>A","T>C","T>G"),each=16),rep("INS",4),rep("DEL",4))
    df_lower$Base = c(types,rep(c("A","C","G","T"),2))
    df_upper$Type = c(rep(c("C>A","C>G","C>T","T>A","T>C","T>G"),each=16),rep("INS",4),rep("DEL",4))
    df_upper$Base = c(types,rep(c("A","C","G","T"),2))
    df_lower = melt(df_lower, id.vars = c("Type","Base"))
    df_upper = melt(df_upper, id.vars = c("Type","Base"))
    df3 <- cbind(df2, value_min = df_lower$value, value_max = df_upper$value)
    plot2 = plot + geom_pointrange(data=df3, aes(ymin=value,ymax=value_max,colour = Type_f), fatten = 0.01, size=1, show.legend = F) +
      scale_color_manual(values=c(colors,"white")) +
      geom_pointrange(data=df3, aes(ymin=value_min,ymax=value,colour="white"), size=1, fatten = 0.01,show.legend = F)
  
  return(plot2)
}
```

`plot_96_profile_CI` - function for plotting 96-long mutational profiles of different mutation types with confidence intervals.
Requires: `ggplot2` and `reshape2` libraries.
Arguments:
* `mut_matrix` - matrix of profiles, samples x mutation counts
* `mut_matrix_lower` - matrix of lower CI values for signatures, mutation types x samples
* `mut_matrix_upper` - matrix of upper CI values for signatures, mutation types x samples
* `colors` - colors to use for 6 mutation classes
* `size` - size of mutation class titles
* `ymax` - maximal value on the plot
* `CI` - `FALSE` or `TRUE`, do not / do plot CIs
```{r,eval=F}
plot_96_profile_CI <- function (mut_matrix, mut_matrix_lower = NULL, 
                                mut_matrix_upper = NULL, CI=FALSE, 
                                colors=c("deepskyblue2","black","red3","grey","olivedrab3","pink"), ymax=NA, size=9) 
{
  if(is.na(ymax)) ymax = max(mut_matrix_upper)
  if (ymax>1) step = 10
  if (ymax<=1) step = 0.1
  C_TRIPLETS = c(
    "ACA", "ACC", "ACG", "ACT",
    "CCA", "CCC", "CCG", "CCT",
    "GCA", "GCC", "GCG", "GCT",
    "TCA", "TCC", "TCG", "TCT")
  T_TRIPLETS = c(
    "ATA", "ATC", "ATG", "ATT",
    "CTA", "CTC", "CTG", "CTT",
    "GTA", "GTC", "GTG", "GTT",
    "TTA", "TTC", "TTG", "TTT")
  types <- c(rep(C_TRIPLETS,3), rep(T_TRIPLETS,3))
  context = types
  substitution = rep(c("C>A","C>G","C>T","T>A","T>C","T>G"), each = 16)
  substring(context, 2, 2) = "*"
  df = data.frame(substitution = substitution, context = context)
  df2 = cbind(df, as.data.frame(mut_matrix))
  df3 = melt(df2, id.vars = c("substitution", "context"))
  value = NULL
  plot = ggplot(data = df3, aes(x = context, y = value, fill = substitution, width = 0.6)) + 
    geom_bar(stat = "identity", colour = "black", size = 0.2) + 
    scale_fill_manual(values = colors) + 
    facet_grid(variable ~ substitution) + 
    coord_cartesian(ylim = c(0,ymax)) + 
    scale_y_continuous(breaks = seq(0, ymax, step)) +
    guides(fill = FALSE) + theme_bw() + 
    theme(text = element_text(family='ArialMT'),
          axis.title.y = element_text(size = 14,vjust = 1), 
          axis.text.y = element_text(size = 8), 
          axis.title.x = element_text(size = 14), 
          axis.text.x = element_text(size = 8, angle = 90, vjust = 0.4), 
          strip.text.x = element_text(size = 16), 
          strip.text.y = element_text(size = size), 
          panel.grid.major.x = element_blank(), 
          strip.background = element_blank(), 
          panel.border = element_rect(colour="white"),
          panel.spacing = unit(0.1,'lines')) +
    labs(y="Number of mutations", x="Context")
  
  if (CI) { # add confidence intervals
    rownames(mut_matrix_upper) = rownames(mut_matrix_lower) = NULL
    df_lower = cbind(df, as.data.frame(mut_matrix_lower))
    df_upper = cbind(df, as.data.frame(mut_matrix_upper))
    df_lower = melt(df_lower, id.vars = c("substitution", "context"))
    df_upper = melt(df_upper, id.vars = c("substitution", "context"))
    df2 = cbind(df, as.data.frame(mut_matrix[,colnames(mut_matrix_lower)]))
    df3 = melt(df2, id.vars = c("substitution", "context"))
    df3 <- cbind(df3, value_min = df_lower$value, value_max = df_upper$value)
    plot = plot + geom_pointrange(data=df3, aes(ymin=value,ymax=value_max,colour = substitution), size=1.5, fatten = 0.001,show.legend = F) +
      scale_color_manual(values=c(colors,"white")) +
      geom_pointrange(data=df3, aes(ymin=value_min,ymax=value,colour="white"), size=1.5, fatten = 0.001,show.legend = F)
  }
  return(plot)
}
```


### Functions for NMF: extracting signatures, calculating contributions.

`nmSolve` - extraction of signatures for given exposures.
Arguments:
* `D` - matrix of mutation counts, samples x mutation types
* `P` - exposures, samples x signatures
* `maxIter` - maximal number of iterations in NMF algorithm
* `tol` - lower bound for the change of signature matrix between iterations
* `div.err` - lower bound for the change of the divergence between D and its factorization between iterations
```{r}
nmSolve <- function(D, P, maxIter = 10000, tol=1e-5, div.err=1e-7) {
  n <- nrow(D)
  mask <- !is.na(D)
  m <- ncol(D)
  s <- ncol(P)
  rP <- rep(colSums(P), m)
  tP <- t(P)
  D <- as.matrix(D)
  P <- as.matrix(P)
  E1 <- E2 <- matrix(runif(s * m, 1e-3, 1), ncol = m)
  err <- 2*tol
  D[is.na(D)] <- 0
  
  iter <- 1
  divergence.old <- mean(D*log(D/(P %*% (E2 + .Machine$double.eps))) - D + P%*%E2, na.rm=T)
  div.change <- 2 * div.err
  
  while (iter < maxIter & err > tol & abs(div.change) > div.err) {
    E1 <- E2
    E2 <- E1 * (tP %*% ((mask*D)/(mask*(P %*% (E1 + .Machine$double.eps)) + .Machine$double.eps)))/rP
    iter <- iter + 1
    err <- mean(abs(E2 - E1)/(E1+.Machine$double.eps), na.rm=TRUE)
    divergence <- mean(D*log(D/(P %*% (E2 + .Machine$double.eps))) - D + P%*%E2, na.rm=T) # KL distance from D to P%*%E2
    div.change <- divergence.old - divergence
    divergence.old = divergence
    if(iter %% 100 == 0) cat(round(-log10(err)))
  }
  cat("\n")
  if(iter == maxIter) warning(paste("No convergence after",iter, "iterations."))
  E2
}
```

`nmFit` - fitting the signatures (extracting exposures).
Arguments:
* `D` - matrix of mutation counts, samples x mutation types
* `E` - signatures, signatures x mutation types
* `maxIter` - maximal number of iterations in NMF algorithm
* `tol` - lower bound for the change of exposure matrix between iterations
* `div.err` - lower bound for the change of the divergence between D and its factorization between iterations
```{r}
nmFit <- function(D, E, maxIter = 10000, tol=1e-5, div.err=1e-7) {
  n <- nrow(D)
  m <- ncol(D)
  s <- nrow(E)
  tE <- t(E)
  rE <- rep(rowSums(E),each=n)
  D <- as.matrix(D)
  E <- as.matrix(E)
  P1 <- P2 <- matrix(runif(s * n, 1e-3, 1), nrow = n)
  err <- 2*tol
  
  iter <- 1
  divergence.old <- mean(D * log(D/(P2%*%(E + .Machine$double.eps))) - D + P2%*%E, na.rm=T)
  div.change <- 2 * div.err
  while (iter < maxIter & err > tol & abs(div.change) > div.err) {
    P1 <- P2
    P2 <- P1 * ((D/((P1 +.Machine$double.eps) %*% (E +.Machine$double.eps))) %*% tE) / (rE+.Machine$double.eps)
    iter <- iter + 1
    err <- mean(abs(P2 - P1)/(P1+.Machine$double.eps), na.rm=TRUE)
    divergence <- mean(D*log(D/((P2 + .Machine$double.eps)%*%(E+.Machine$double.eps))) - D + P2%*%E, na.rm=T) # KL distance from D to P%*%E
    div.change <- divergence.old - divergence
    divergence.old = divergence
    if(iter %% 100 == 0) cat(round(-log10(err)))
    # add likelihood convergence
  }
  cat("\n")
  if(iter == maxIter) warning(paste("No convergence after",iter, "iterations."))
  P2
}
```


### Other

Technical function: slightly modified function from `mg14` to plot cirles with different rim colour.
```{r, eval=F}
corr_scatterpie <- function (x, y, p, r, xlab = "", ylab = "", circles = FALSE, 
          lwd.circle = rep(1, length(x)), lty.circle = rep(1, length(x)), 
          add = FALSE, col.circle='black', ...) 
{
  if (!add) 
    plot(x, y, xlab = xlab, ylab = ylab, pch = NA)
  for (i in seq_along(x)) {
    mg14:::.pie(p[i, ], x0 = x[i], y0 = y[i], radius = r[i], add = TRUE, 
         ...)
    if (circles) {
      u <- par("usr")
      pr <- (u[2] - u[1])/(u[4] - u[3])
      fr <- par("pin")[1]/par("pin")[2]
      polygon(x[i] + cos(seq(0, 2 * pi, l = 100)) * r[i], 
              y[i] + sin(seq(0, 2 * pi, l = 100)) * r[i]/pr * 
                fr, col = NA, lty = lty.circle[i], lwd = lwd.circle[i], border=col.circle)
    }
  }
}
```


Get COSMIC signatures

```{r, warning=F, eval=F}
sp_url <- paste("http://cancer.sanger.ac.uk/cancergenome/assets/",
                "signatures_probabilities.txt", sep = "")
cancer_signatures = read.table(sp_url, sep = "\t", header = TRUE)
cancer_signatures = cancer_signatures[order(cancer_signatures[,1]),]
types <- as.character(cancer_signatures$Trinucleotide) # trinucleotide classes
types.full <- as.character(cancer_signatures$Somatic.Mutation.Type) # substitution types
row.names(cancer_signatures) <- types.full
cancer_signatures = as.matrix(cancer_signatures[,4:33])
```

Adjust them from exomes:

```{r, warning=F, eval=F}
cancer_signatures_adj <- cancer_signatures
for (i in 1:ncol(cancer_signatures)) {
  cancer_signatures_adj[,i] <- cancer_signatures_adj[,i] / tri.counts.genome[types,1] * tri.counts.exome[types,1]
  cancer_signatures_adj[,i] <- cancer_signatures_adj[,i] / sum(cancer_signatures_adj[,i])
}
```

```{r, echo=F, fig.width=10, fig.height=4, eval=F}
to.show <- cbind(cancer_signatures[,5], cancer_signatures_adj[,5])
colnames(to.show) <- c('Sig.5 genome', 'Sig.5 exome')
plot_sig_wb(to.show)
```

Prepare human exome counts (regions taken from [Agilent SureSelect V5 Human All Exon](https://earray.chem.agilent.com/suredesign/index.htm), human genome sequence - from hg19 build)

```{r, eval=F, warning=F, message=F, eval=F}
ref_genome="BSgenome.Hsapiens.UCSC.hg19"
library(BSgenome.Hsapiens.UCSC.hg19)
exome <- read.table("S04380110_Covered.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE, skip=2,quote="")
a <- getSeq(get(ref_genome))
a <- a[1:23] # get rid of Y
gr <- as(seqinfo(a), "GRanges") # turn into GRanges object
genome(gr) <- "hg19"
# Get the sequence for well covered exome
exactexomelist <- list()
for (j in 1:23) {
  tmp <- exome[exome$V1==seqlevels(a)[j],]
  exactexomelist[[j]] <- lapply(1:nrow(tmp), function(i)
    a[[j]][tmp$V2[i]:tmp$V3[i]])
}
```

```{r, echo=F, warning=F, message=F}
load('exome.RData')
# nucleotide counts
exome.bases <- rowSums(sapply(1:23, function(k) rowSums(sapply(exactexomelist[[k]],oligonucleotideFrequency,width=1))))
ref_genome="BSgenome.Hsapiens.UCSC.hg19"
library(BSgenome.Hsapiens.UCSC.hg19)
#exome <- read.table("S04380110_Covered.bed",header = FALSE, sep="\t",stringsAsFactors=FALSE, skip=2,quote="")
a <- getSeq(get(ref_genome))
a <- a[1:23] # get rid of Y
gr <- as(seqinfo(a), "GRanges") # turn into GRanges object
genome(gr) <- "hg19"
```

Prepare *C. elegans* counts in order to account for trinucleotide content difference:

```{r, warning=F, eval=F}
WBcel235 <- readDNAStringSet("~/Desktop/C. elegans WB235/Caenorhabditis_elegans.WBcel235.dna_sm.toplevel.fa.gz")
worm.trinucleotides <- colSums(trinucleotideFrequency(WBcel235)[-5,])
human.trinucleotides <- as.vector(t(tri.counts.genome))
names(human.trinucleotides) <- row.names(tri.counts.genome)
trinucleotide.freq.factor <- sapply(unique(types), function(x) {
  freq.worm <- worm.trinucleotides[x] + worm.trinucleotides[as.character(reverseComplement(DNAString(x)))]
  return(freq.worm /  human.trinucleotides[x]) # tri.counts.genome is already classified w.r.t. pyrimidine reference
})
human.trinucleotides <- as.vector(t(tri.counts.exome)) # / sum(tri.counts.genome))) # counts from "deconstructSigs" package
names(human.trinucleotides) <- row.names(tri.counts.exome)
trinucleotide.freq.factor.ex <- sapply(unique(types), function(x) {
  freq.worm <- worm.trinucleotides[x] + worm.trinucleotides[as.character(reverseComplement(DNAString(x)))]
  return(freq.worm / human.trinucleotides[x]) # tri.counts.genome is already classified w.r.t. pyrimidine reference
})
names(trinucleotide.freq.factor.ex) = names(trinucleotide.freq.factor) <- unique(types)
```

Visualize trinucleotide differences between C.elegans and human exome:

```{r, echo=F, warning=F, fig.width=10, fig.height=4, eval=F}
worm.trinucleotides.32 <- sapply(unique(types), function(x) {
  return(worm.trinucleotides[x] + worm.trinucleotides[as.character(reverseComplement(DNAString(x)))])
})
names(worm.trinucleotides.32) <- unique(types)
types.order <- c(rep(types[1:16],3),rep(types[49:64],3))
# human.trinucleotides in deconstructSigs

df = data.frame(worm.counts=c(worm.trinucleotides.32[types.order]),
                human.counts=c(t(tri.counts.exome)[1,][types.order]),
                Type = rep(c('C>A', 'C>G', 'C>T', 'T>A', 'T>C', 'T>G'), each = 16),
                Trinucleotide = c(types.order))
df$worm.counts <- df$worm.counts / sum(df$worm.counts[33:64])
df$human.counts = df$human.counts / sum(df$human.counts[33:64])
df$human.counts = -df$human.counts

df <- rbind(df, df[,c(2,1,3,4)])
df[97:192,1] <- -df[97:192,1]
df <- df[,-2]
df$Species <- c(rep("C. elegans",96), rep("Homo sapiens exome",96))
colnames(df) <- c('Fraction', 'Type', 'Trinucleotide', 'Species')
rownames(df) = NULL

p <- ggplot(data = df,aes(x = Trinucleotide,y=Fraction,fill=Species)) + 
  geom_bar(stat="identity",colour="black",position = "dodge",size=0.1,width = 0.5) + 
  scale_fill_manual(values = c("darkred","lightblue")) +
  facet_grid(Species ~ Type, scales = "free") +
  scale_x_discrete(labels=types.order) +
  theme_bw() + coord_cartesian() +
  theme(text = element_text(family='ArialMT'),
        axis.text=element_text(size=12),
        axis.text.x = element_text(angle = 90, vjust = 0.4,size=6),
        strip.text = element_text(size = 12),
        axis.title=element_text(size=12,face="bold"),
        legend.text = element_text(size=12),
        legend.title = element_text(size=24,face="bold"),
        panel.grid = element_blank(),
        strip.background = element_rect(colour='white', fill='white'),
        panel.border = element_rect(colour='black', size=0.1),
        panel.spacing = unit(0.01,'lines')) +
  guides(fill=F)
p
```

## *C. elegans* MMR mutational patterns
Upload the mutation counts from *C. elegans* samples and calculate *C. elegans* signatures using additive Poisson model:

$Y_{i,j} = Pois(\lambda_{i,j})$,

$E[Y_{i,j}] = N \cdot (\beta_{j,b} + X_{g_{1}} \beta_{j,g_{1}} + X_{g_{2}} \beta_{j,g_{2}} + X_{g_{1}:g_{2}} \beta_{j,g_{1}:g_{2}})$, 

where $β_{j,\cdot} \ge 0$ - effects, $N$ - generation number, $g_{1}$, $g_{2}$ - genetic backgrounds, $b$ - background contribution, $X_{...} \in {0,1}$ indicates the presence of particular factors. 

```{r, warning=F, message=F, fig.width=10, fig.height=6, results='hide', eval=F}
load("~/Desktop/Git/MMR_paper_analysis_final/C. elegans data/Learned_signatures.RData")
# Contains mutation counts matrix mut_mat and exposure matrix small.X for the samples from C. elegans MMR and pole-4;pms-2 experiments.
learned.sigs <- nmSolve(t(mut_mat),small.X,maxIter=10000, tol = 1e-06, div.err = 1e-10)
for (i in 1:nrow(learned.sigs)) {
  learned.sigs[i,] <- learned.sigs[i,] / sum(learned.sigs[i,])
}
```

```{r, warning=F, message=F, fig.width=10, fig.height=6, eval=F}
plot_sig_wb(t(learned.sigs))
```

Humanize the signatures:

```{r, warning=F, fig.width=10, fig.height=6, eval=F}
learned.sigs.exome <- learned.sigs
for (i in 1:nrow(learned.sigs.exome)) {
  learned.sigs.exome[i,] <- learned.sigs.exome[i,] / trinucleotide.freq.factor.ex[types]
}
plot_sig(t(learned.sigs.exome))
```

## ICGC data

ICGC data for COAD and STAD dataset was downloaded from [ICGC DCC](http://dcc.icgc.org) using COAD-US and STAD-US projects with WXS analysis type, respectively. The vcf file describing all somatic mutations across ICGC dataset is stored [here](https://dcc.icgc.org/releases/current/Summary) under the 'simple_somatic_mutation.aggregated.vcf.gz' name.

```{r, eval=F, warning=F, eval=F}
# download and read huge aggregated variant file
big.vcf <- readVcf("simple_somatic_mutation.aggregated.vcf")
mutations.COAD <- read.table(file="COAD/simple_somatic_mutation.open.tsv",sep="\t",header=T)
mutations.STAD <- read.table(file="STAD/simple_somatic_mutation.open.tsv",sep="\t",header=T)
per_sample_list <- sapply(unique(mutations.COAD$icgc_donor_id), function(donor) {
  unique(mutations.COAD$icgc_mutation_id[mutations.COAD$icgc_donor_id==donor])
})
names(per_sample_list) <- unique(mutations.COAD$icgc_donor_id)
vcf_list_COAD <- sapply(per_sample_list, function(x) big.vcf[as.character(x)])
names(vcf_list_COAD) <- names(per_sample_list)
vcf_list_COAD <- sapply(vcf_list_COAD, rowRanges)
for (sample in names(vcf_list_COAD)) {
  seqlevels(vcf_list_COAD[[sample]]) <- seqnames(get(ref_genome))[c(as.numeric(seqlevels(vcf_list_COAD[[sample]])[1:22]),25,23,24)]
  vcf_list_COAD[[sample]] <- vcf_list_COAD[[sample]][seqnames(vcf_list_COAD[[sample]])!="chrM"]
  vcf_list_COAD[[sample]] <- vcf_list_COAD[[sample]][seqnames(vcf_list_COAD[[sample]])!="chrY"]
}
per_sample_list <- sapply(unique(mutations.STAD$icgc_donor_id), function(donor) {
  unique(mutations.STAD$icgc_mutation_id[mutations.STAD$icgc_donor_id==donor])
})
names(per_sample_list) <- unique(mutations.STAD$icgc_donor_id)
vcf_list_STAD <- sapply(per_sample_list, function(x) big.vcf[as.character(x)])
names(vcf_list_STAD) <- names(per_sample_list)
vcf_list_STAD <- sapply(vcf_list_STAD, rowRanges)
for (sample in names(vcf_list_STAD)) {
  seqlevels(vcf_list_STAD[[sample]]) <- seqnames(get(ref_genome))[c(as.numeric(seqlevels(vcf_list_STAD[[sample]])[1:22]),25,23,24)]
  vcf_list_STAD[[sample]] <- vcf_list_STAD[[sample]][seqnames(vcf_list_STAD[[sample]])!="chrM"]
  vcf_list_STAD[[sample]] <- vcf_list_STAD[[sample]][seqnames(vcf_list_STAD[[sample]])!="chrY"]
}

all.types <- c(types.full,"INS_A","INS_C","INS_G","INS_T","DEL_A","DEL_C","DEL_G","DEL_T")
sub_list <- sapply(vcf_list_COAD, function(vcf) {
  vcf[width(vcf$REF)==1 & width(unlist(vcf$ALT))==1,]
})
del_list <- sapply(vcf_list_COAD, function(vcf) {
  vcf[width(vcf$REF)==2 & width(unlist(vcf$ALT))==1,]
})
ins_list <- sapply(vcf_list_COAD, function(vcf) {
  vcf[width(vcf$REF)==1 & width(unlist(vcf$ALT))==2,]
})
COAD.mutation.counts = matrix(0,nrow=length(vcf_list_COAD),ncol=length(all.types),dimnames=list(names(vcf_list_COAD),all.types))
for (i in 1:nrow(COAD.mutation.counts)) {
  type_context = type_context(sub_list[[i]], ref_genome)
  counts <- table(type_context)
  for (a in rownames(counts)) {
    tmp = unlist(strsplit(a,split="[>]"))
    inds <- colnames(counts)[counts[a,]>0]
    columns = as.vector(sapply(inds, function(x) paste(substr(x,1,1),"[",a,"]",substr(x,nchar(x),nchar(x)),sep="")))
    COAD.mutation.counts[i,columns] = counts[a,inds]
  }
  COAD.mutation.counts[i,97:100] <- table(substr(unlist(ins_list[[i]]$ALT),2,2))[c("A","C","G","T")]
  COAD.mutation.counts[i,101:104] <- table(substr(del_list[[i]]$REF,2,2))[c("A","C","G","T")]
  print(i)
}
COAD.mutation.counts[is.na(COAD.mutation.counts)] <- 0

sub_list <- sapply(vcf_list_STAD, function(vcf) {
  vcf[width(vcf$REF)==1 & width(unlist(vcf$ALT))==1,]
})
del_list <- sapply(vcf_list_STAD, function(vcf) {
  vcf[width(vcf$REF)==2 & width(unlist(vcf$ALT))==1,]
})
ins_list <- sapply(vcf_list_STAD, function(vcf) {
  vcf[width(vcf$REF)==1 & width(unlist(vcf$ALT))==2,]
})
STAD.mutation.counts = matrix(0,nrow=length(vcf_list_STAD),ncol=length(all.types),dimnames=list(names(vcf_list_STAD),all.types))
for (i in 1:nrow(STAD.mutation.counts)) {
  type_context = type_context(sub_list[[i]], ref_genome)
  counts <- table(type_context)
  for (a in rownames(counts)) {
    tmp = unlist(strsplit(a,split="[>]"))
    inds <- colnames(counts)[counts[a,]>0]
    columns = as.vector(sapply(inds, function(x) paste(substr(x,1,1),"[",a,"]",substr(x,nchar(x),nchar(x)),sep="")))
    STAD.mutation.counts[i,columns] = counts[a,inds]
  }
  STAD.mutation.counts[i,97:100] <- table(substr(unlist(ins_list[[i]]$ALT),2,2))[c("A","C","G","T")]
  STAD.mutation.counts[i,101:104] <- table(substr(del_list[[i]]$REF,2,2))[c("A","C","G","T")]
  print(i)
}
STAD.mutation.counts[is.na(STAD.mutation.counts)] <- 0
```

Or just upload the prepared data:

```{r, warning=F, eval=F}
load("ICGC data/profiles_and_decomposition.RData")
```

# Signature extraction

Signature extraction from the whole set is performed via Brunet version non-negative matrix factorization using NMF package. The number of signatures is chosen as the number of signatures where both residual sum of squares and Akakike Information Criterion values stabilize.

```{r, echo=F, warning=F, eval=F}
divergence <- function (a,b) {
  return (a * log ( (a+.Machine$double.eps)/(b + .Machine$double.eps)) - a + b)
}
mm <- rbind(COAD.mutation.counts,STAD.mutation.counts)
mut_mat = t(mm) + 0.0001
rss <- NULL; likelihoods <- NULL; divergences <- NULL
for (r in 2:12) {
  res <- NMF::nmf(x=mut_mat,rank=r,seed=123456,method='brunet')
  sigs <- NMF::basis(res)
  cont <- NMF::coef(res)
  rss <- c(rss, sum((t(mm) - sigs%*%cont)**2))
  likelihoods = c(likelihoods,sum(dpois(x=t(mm),lambda=sigs%*%cont,log=T)))
  divergences <- c(divergences,sum(divergence(t(mm),sigs%*%cont)))
}
AIC = 2*c(2:12)*(ncol(mm) + nrow(mm)) - 2*likelihoods
```

Plot AIC:

```{r, warning=F, fig.width=6, fig.height=3}
df = data.frame(rank=2:12,AIC)
ggplot(df,aes(x=rank,y=AIC)) + geom_point() + 
  ggtitle("AIC for selecting the number of signatures") + 
  geom_hline(yintercept = AIC[7],linetype = "longdash",colour="red")
```

Now plot RSS per rank:

```{r, warning=F, fig.width=6, fig.height=3, eval=F}
df = data.frame(rank=2:12,rss)
ggplot(df,aes(x=rank,y=rss)) + geom_point() + 
  ggtitle("RSS for selecting the number of signatures") + 
  geom_hline(yintercept = rss[7],linetype = "longdash",colour="red")
```

This is how the final set of signatures looks like:

```{r, warning=F, fig.width=10, fig.height=8, eval=F}
res <- NMF::nmf(x=mut_mat,rank=8,seed=123456,method='brunet')
sigs <- NMF::basis(res)
sigs <- sigs[,c(8,1,2,5,4,7,3,6)] # Reorder the signatures
decomposition <- t(NMF::coef(res))
decomposition <- decomposition[,c(8,1,2,5,4,7,3,6)]
colnames(sigs) <- c("Clock-1", "Clock-2", "POLE", "17-like", "MMR-1", "MMR-2", "MMR-3", "SNP")
colnames(decomposition) <- colnames(sigs)
for (i in 1:nrow(decomposition))
   decomposition[i,] = decomposition[i,]/sum(decomposition[i,])
plot_sig_104(sigs,size=12)
```

Signatures are assigned as follows: Clock-1 (5meC), Clock-2 (+APOBEC), POLE, 17-like, MMR-1 (20), MMR-2 (15), MMR-3 (21), MMR-4 (unknown).

Get microsatellite stability/instability (MSS/MSI status) from UCSC (or rather the Cancer Genome Atlas Clinical Explorer):

```{r, warning=F, eval=F}
icgc.coad <- rownames(COAD.mutation.counts)
icgc.stad <- rownames(STAD.mutation.counts)
coad <- read.delim('~/Desktop/Git/MMR_paper_analysis_final/ICGC data/COAD/donor.tsv',header=T)
stad <- read.delim('~/Desktop/Git/MMR_paper_analysis_final/ICGC data/STAD/donor.tsv',header=T)
tcga.coad <- as.character(coad$submitted_donor_id[match(icgc.coad, coad$icgc_donor_id)])
tcga.stad <- as.character(stad$submitted_donor_id[match(icgc.stad, stad$icgc_donor_id)])

mss_coadread <- read.delim('~/Desktop/Git/MMR_paper_analysis_final/ICGC data/COAD/COADREAD_2015-04-02_ClinicalParameters.txt',header=T)
mss_stad <- read.delim('~/Desktop/Git/MMR_paper_analysis_final/ICGC data/STAD/STAD_2015-04-02_ClinicalParameters.txt',header=T)
mmr.coad <- icgc.coad[match(intersect(mss_coadread$SampleCode[mss_coadread$MSIstatus=='MSI-H'], tcga.coad),tcga.coad)]
mmr.stad <- icgc.stad[match(intersect(mss_stad$SampleCode[mss_stad$MSIstatus=='MSI-H'], tcga.stad),tcga.stad)]
mmr.ucsc <- c(mmr.stad, mmr.coad)
```

To assess the distribution of signatures in the data, we generate tSNE plot based on cosine similarity between sample profiles. Circle sizes reflect number of mutations, black rim - microsatellite instability (MSI high), coloured sectors correspond to relative contributions of different signatures. Note that all MSI samples group in one cluster across both datasets.

```{r, echo=F, message=FALSE, warning=FALSE, errors = FALSE, eval=F}
data <- rbind(COAD.mutation.counts, STAD.mutation.counts)
w <- which(rowSums(data)>0)
data <- data[w,]
cosdist <- function(x,y) {
	x0 <- x/sum(x)
	y0 <- y/sum(y)
	x0 %*% y0 / sqrt(x0%*%x0)/sqrt(y0%*%y0)
}
D <- as.dist(sapply(1:nrow(data), function(i) sapply(1:nrow(data), function(j) 1-cosdist(data[i,],data[j,]) )))
set.seed(1)
t <- tsne(D)
MMR = intersect(rownames(data), mmr.ucsc)
col <- c(RColorBrewer::brewer.pal(8,"Set1")[c(1:2,7:8,3:5)],'darkgrey')
m <- rownames(data) %in% MMR
```

Visualize the similarity map:

```{r, message=FALSE, warning=FALSE, errors = FALSE, echo=F, eval=F}
# Plot everything
o1 <- order(rowSums(data[!m,]),decreasing = T)
o2 <- order(rowSums(data[m,]),decreasing = T)
par(bty="n", mar=c(0,0,0,0))
plot(NA,NA, xlab="", ylab="", xlim=c(-20,20), ylim=c(-35,35), xaxt="n", yaxt="n")
corr_scatterpie(t[!m,1][o1], t[!m,2][o1], p=decomposition[w,][!m,][o1,], r=sqrt(rowSums(data)[!m][o1])/75, labels=NA, col=col, lty=0, circles=TRUE, lwd.circle=rep(0.01,sum(!m)),lty.circle=rep(1,sum(!m)), add=TRUE, col.circle = 'white')
corr_scatterpie(t[m,1][o2], t[m,2][o2], p=decomposition[w,][m,][o2,], r=sqrt(rowSums(data)[m][o2])/75, labels=NA, col=col, lty=0, circles=TRUE, lwd.circle=rep(2.5,sum(m)), lty.circle=rep(1,sum(m)), add=TRUE, col.circle = 'black')
par(cex=0.66)
mg14:::.pie(x0=10, y0=27, x=matrix(rep(1,8), nrow=1), r=sqrt(10000)/75, labels=colnames(decomposition), col=col, lty=0, circles=TRUE, add=TRUE)
u <- par("usr")
pr <- (u[2]-u[1])/(u[4]-u[3])
fr <- par("pin")[1]/par("pin")[2]
for(i in c(1,10,100,1000,10000)){
	polygon(20 + cos(seq(0,2*pi, l=100)) * sqrt(i)/75, 25+(1+sin(seq(0,2*pi, l=100))) * sqrt(i)/75 / pr * fr, col=NA)
  if (i>10) text(20, 25 + 2*sqrt(i)/75 / pr * fr + 0.3,labels = as.character(i),cex=0.8)
}
polygon(20 + cos(seq(0,2*pi, l=100)) * sqrt(i)/75, 15+(1+sin(seq(0,2*pi, l=100))) * sqrt(i)/75 / pr * fr, lwd=2.5, col=NA)
text(x = 20,y=24,labels = "Number of mutations")
text(x = 20,y=13,labels = "MSI-H status")
```

## MMR

Let's plot signature decomposition for MSI samples in both datasets:

```{r, warning=F, echo=F, fig.width=10, fig.height=6, eval=F}
p2 <- plot_decomposition(decomposition,mm,intnames=intersect(mmr.ucsc,rownames(STAD.mutation.counts)),col=col) + ggtitle("STAD MSI samples")
p1 <- plot_decomposition(decomposition,mm,intnames=intersect(mmr.ucsc,rownames(COAD.mutation.counts)),col=col) + ggtitle("COAD MSI samples")
multiplot(p1,p2)
```

Association with MSI status: calculate P-values for relative signature contributions compared with MSI status.

```{r, warning=F, eval=F}
pvals <- NULL
for (i in 1:8)
{
  pvals <- c(pvals, wilcox.test(decomposition[mmr.ucsc,i], decomposition[setdiff(rownames(mm),mmr.ucsc),i], paired=F,alternative='greater')$p.value)
}
names(pvals) = colnames(decomposition)
print(p.adjust(pvals,method = 'bonf')) # Take 0.01 as significance threshold (>10 tests)
```

MMR-1 seems to be a good indicator of MSI status. Calculate AUC for MMR-1:

```{r, warning=F, message=F, eval=F}
library(pROC)
mmr.type <- sapply(rownames(mm), function(x) {
  if (x %in% mmr.ucsc) return("MSI")
  else return("MSS")
})
roc(response = factor(mmr.type), predictor = decomposition[,5])
ci(response = factor(mmr.type), predictor = decomposition[,5])
```

Boxplots for signature MMR-1, MMR-2, MMR-3 contribution vs MSI status:

```{r, warning=F, echo=F, results='hide', eval=F}
df = data.frame(sample = rownames(mm), mmr.type, decomposition[,c(5:7)])
colnames(df)[3:5] = colnames(decomposition)[c(5:7)]
df = melt(df)
```
```{r,fig.width=6, fig.height=3, echo=F, warning=F, eval=F}
p1 <- ggplot(df, aes(x=mmr.type,y=value,fill=variable)) + geom_boxplot(width=0.4,size=0.1,outlier.size=0.5) +
  facet_grid(. ~ variable) + scale_fill_manual(values=col[c(5:7)]) +
  ylab("Relative contribution") + xlab("") + guides(fill=F) + theme_bw() +
  theme(text = element_text(family='ArialMT'),
        strip.text.x = element_text(size=16),
        axis.title.y = element_text(size=16),
        axis.text.x = element_text(size=10),
        axis.text.y = element_text(size=10),
        panel.grid = element_blank(),
        panel.border = element_rect(size = 0.1),
        strip.background = element_blank())
p1
```

As we can see from the signature plot, all indels are coming with MMR-1; to confirm it, plot the absolute contribution of MMR signatures vs the number of indels:

```{r, warning=F, echo=F, fig.width=4, fig.height=2, eval=F}
new.cont.mat <- t(decomposition)
for (i in 1:ncol(new.cont.mat))
  new.cont.mat[,i] <- new.cont.mat[,i] * rowSums(mm)[i]
df = data.frame(sample=rownames(mm),ind=rowSums(mm[,97:104]),mmr1=new.cont.mat[5,],mmr2=new.cont.mat[6,],mmr3=new.cont.mat[7,])
df <- melt(df, id=c('sample','ind'))
p <- ggplot() + geom_point(data = df[df$variable=='mmr3',], aes(x=ind,y=value), col=col[7],size=0.5) +
  geom_point(data = df[df$variable=='mmr2',], aes(x=ind,y=value), col=col[6],size=0.5) + 
  geom_point(data = df[df$variable=='mmr1',], aes(x=ind,y=value), col=col[5],size=0.5) +
  theme_bw() +
  theme(text = element_text(family='ArialMT'),
        strip.text.x = element_text(size=16),
        axis.title.y = element_text(size=16),
        axis.text.x = element_text(size=10),
        axis.text.y = element_text(size=10),
        panel.grid = element_blank(),
        panel.border = element_rect(size = 0.1),
        strip.background = element_blank()) 
  #xlab('Number of 1 bp indels') + ylab('Number of mutations assigned to a signature') + 
p
```

To see the unusual contribution of Clock-1 signature, plot the average fold change (log scale) in the number of mutations assigned to different signatures in MSI samples and MSS samples:

```{r, warning=F, echo=F, fig.width=8, fig.height=2, eval=F}
df = data.frame(sigs = colnames(decomposition), val=log10(rowMeans(new.cont.mat[,mmr.ucsc]) / rowMeans(new.cont.mat[,setdiff(rownames(mm),mmr.ucsc)])))
p <- ggplot(data=df, aes(x=factor(sigs,levels=rev(colnames(decomposition))),y=val)) + 
  geom_bar(stat='identity', fill=rev(col)[c(6,1:5,7:8)]) + 
  coord_flip() + 
  theme_bw() +
  theme(text = element_text(family='ArialMT'),
        strip.text.x = element_text(size=16),
        axis.title.y = element_text(size=16),
        axis.text.x = element_text(size=10),
        axis.text.y = element_text(size=10),
        panel.grid = element_blank(),
        panel.border = element_rect(size = 0.1),
        strip.background = element_blank()) +
  ylab('no. mutations in MSI / no. mutations in MSS') + xlab('Signatures') +
  scale_y_continuous(breaks=c(0,1,2),labels=c(1,10,100))
p
```


Relative contributions for MMR deficient samples:
```{r, warning=F, echo=F, fig.width=10, fig.height=6, eval=F}
p1 <- plot_decomposition(decomposition,decomposition,intersect(rownames(COAD.mutation.counts),mmr.ucsc),col,axis.size=20)
p2 <- plot_decomposition(decomposition,decomposition,intersect(rownames(STAD.mutation.counts),mmr.ucsc),col,size=4, axis.size=20)
multiplot(p1,p2)
```

Absolute contributions for all the samples in both datasets:

```{r, warning=F, echo=F, fig.width=10, fig.height=6, eval=F}
p1 <- plot_decomposition(decomposition,mm,rownames(COAD.mutation.counts),col,size=2, axis.size=14)
p2 <- plot_decomposition(decomposition,mm,rownames(STAD.mutation.counts),col, size=2, axis.size=14)
multiplot(p1,p2)
```

## Amount of indels 

Plot the amount of indels per sample in both cohorts:

```{r, warning=F, echo=F, fig.width=10, fig.height=4, eval=F}
indels <- rowSums(mm[,97:104])
cohort <- ifelse(rownames(mm)[indels>10] %in% rownames(COAD.mutation.counts),"COAD","STAD")
df <- data.frame(indelcount=indels[indels>10],cohort,sample=rownames(mm)[indels>10])
ggplot(data=df, aes(x=factor(sample,levels=names(sort(indels))),y=indelcount,fill=cohort)) + geom_bar(stat="identity") + facet_grid(. ~ cohort,scales = "free")
```

Lets find all the homopolymers of length 4 to 55 in human exome.

1. Create homopolymer library

```{r, eval=F, warning=F}
homopolymer_pool_length_4to55 <- lapply(seq(4,55),function(y) lapply(c("A","C","G","T"),function(x)
  if (x=="A") {paste0("B",paste(rep(x,y),collapse=""),"B")
  }
  else
    if (x=="T") {paste0("V",paste(rep(x,y),collapse=""),"V")
    }
  else
    if (x=="G") {paste0("H",paste(rep(x,y),collapse=""),"H")
    }
  else
    if (x=="C") {paste0("D",paste(rep(x,y),collapse=""),"D")
    }
))
homopolymers <- unlist(homopolymer_pool_length_4to55)
pattern_all <- DNAStringSet(homopolymers)
```

2. Find all homopolymers in the genome:

```{r, eval=F, warning=F}
hits_all <- list()
for (i in 1:23) {
  maskMotif(a[[i]], "N") -> masked
  hits_all[[i]] <- sapply(pattern_all, 
                          matchPattern, 
                          subject=masked,
                          fixed=F)
}
lengths <- lapply(hits_all, function(l) sapply(l,length))
hits_all <- lapply(1:length(hits_all), function(i) hits_all[[i]][which(lengths[[i]]>0)])
nonzero_lengths <- lapply(hits_all, function(chr) sapply(chr, length)) # numbers of homopolymers on each chromosome for non empty classes only
names(hits_all) <- seqlevels(gr)
```

3. Create a dataframe withtheir coordinates and contexts:

```{r, eval=F, warning=F}
sites.gr <- lapply(1:23, function(i) do.call("c",lapply(lapply(hits_all[[i]], as, "IRanges"),
                                                        GRanges,seqnames=names(hits_all)[i])))
for (j in 1:23) {
  sites.gr[[j]]$pattern.searched <- rep(as.character(pattern_all), lengths[[j]])
  sites.gr[[j]]$motif.found <- unlist(lapply(hits_all[[j]], as.character))
  sites.gr[[j]]$pattern.length <- unlist(lapply(hits_all[[j]], width)) # includes the two flanking bases
  sites.gr[[j]]$homopolymer.length <- (sites.gr[[j]]$pattern.length)-2 # remove 2 bases, 5' and 3' are not part of homopolymer
}
sites.gr_all <- do.call("c",sites.gr)
genome(sites.gr_all) <- "hg19" # add genome info
```

4. Intersect with well-covereed part of the human exome:

```{r, eval=F, warning=F}
exomranges <- GRanges(seqnames=exome$V1,ranges=IRanges(start = exome$V2,end=exome$V3))
hits <- findOverlaps(exomranges,sites.gr_all,minoverlap = 2)
sites.gr_all <- sites.gr_all[sort(unique(subjectHits(hits)))]
all_sites <- as.data.frame(sites.gr_all) # make a data frame
```

4. Check the amount of 1bp indels in homopolymers in MSI samples in both cohorts:

```{r, warning=F, eval=F}
# Upload the vcf files for both cohorts
load("~/Desktop/Git/MMR_paper_analysis_final/ICGC data/STAD/vcf_list_STAD.RData")
indel_list_STAD <- sapply(vcf_list_STAD, function(vcf) vcf[abs(width(vcf$REF)-width(unlist(vcf$ALT)))==1 & (width(vcf$REF)==1 | width(unlist(vcf$ALT))==1),])
dims <- sapply(indel_list_STAD,length,simplify = T)
load("~/Desktop/Git/MMR_paper_analysis_final/ICGC data/COAD/vcf_list_COAD.RData")
indel_list_COAD <- sapply(vcf_list_COAD, function(vcf) vcf[abs(width(vcf$REF)-width(unlist(vcf$ALT)))==1 & (width(vcf$REF)==1 | width(unlist(vcf$ALT))==1),])
dims <- sapply(indel_list_COAD,length,simplify = T)
for (k in 1:length(indel_list_STAD)) {
  genome(indel_list_STAD[[k]]) <- "hg19"
  indel_list_STAD[[k]]$insertion <- width(unlist(indel_list_STAD[[k]]$ALT))-1
  indel_list_STAD[[k]]$deletion <- width(indel_list_STAD[[k]]$REF)-1
}
for (k in 1:length(indel_list_COAD)) {
  genome(indel_list_COAD[[k]]) <- "hg19"
  indel_list_COAD[[k]]$insertion <- width(unlist(indel_list_COAD[[k]]$ALT))-1
  indel_list_COAD[[k]]$deletion <- width(indel_list_COAD[[k]]$REF)-1
}
indels.in.hp.stad <- vector("numeric",length(indel_list_STAD))
for (i in 1:length(indel_list_STAD)) {
  x <- subsetByOverlaps(indel_list_STAD[[i]],sites.gr_all)
  indels.in.hp.stad[i] <- nrow(as.data.frame(x))
}
names(indels.in.hp.stad) <- names(indel_list_STAD)
indels.in.hp.coad <- vector("numeric",length(indel_list_COAD))
for (i in 1:length(indel_list_COAD)) {
  x <- subsetByOverlaps(indel_list_COAD[[i]],sites.gr_all)
  indels.in.hp.coad[i] <- nrow(as.data.frame(x))
}
names(indels.in.hp.coad) <- names(indel_list_COAD)
```

Plot indels in homopolymers:

```{r, warning=F,fig.width=10, fig.height=4, eval=F}
df <- data.frame(sample=rownames(STAD.mutation.counts), Non.HP.indels = rowSums(STAD.mutation.counts[,97:104])-indels.in.hp.stad[rownames(STAD.mutation.counts)], HP.indels = indels.in.hp.stad[rownames(STAD.mutation.counts)])
df = melt(df,id.vars="sample")
p1 <- ggplot(data=df[df$sample %in% mmr.ucsc,], aes(x=factor(sample, levels = df$sample[order(rowSums(STAD.mutation.counts[,97:104]), decreasing = F)]),y=value,fill=variable)) + geom_bar(stat="identity") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90,vjust=0.5, hjust=0, size=8, family='ArialMT'),
        axis.text.y = element_text(size=8, family='ArialMT'),
        legend.title=element_blank(),
        axis.title = element_text(size=8, family='ArialMT'),
        legend.text = element_text(size=8,family='ArialMT'),
        title = element_text(size=10, family='ArialMT', face='bold'),
        panel.border = element_rect(colour='white'),
        panel.grid = element_blank()) +
  scale_fill_discrete(labels=c("Indels not in HP","Indels in HP")) + xlab("MSI sample") + 
  ylab("1bp indel counts") + ggtitle("1 bp indels in homopolymers in STAD samples with MSI")
p1
```
```{r, warning=F, echo=F, fig.width=10, fig.height=4, eval=F}
df <- data.frame(sample=rownames(COAD.mutation.counts), Non.HP.indels = rowSums(COAD.mutation.counts[,97:104])-indels.in.hp.coad[rownames(COAD.mutation.counts)], HP.indels = indels.in.hp.coad[rownames(COAD.mutation.counts)])
df = melt(df,id.vars="sample")
p2 <- ggplot(data=df[df$sample %in% mmr.ucsc,], aes(x=factor(sample, levels = df$sample[order(rowSums(COAD.mutation.counts[,97:104]), decreasing = F)]),y=value,fill=variable)) + geom_bar(stat="identity") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90,vjust=0.5, hjust=0, size=8, family='ArialMT'),
        axis.text.y = element_text(size=8, family='ArialMT'),
        legend.title=element_blank(),
        axis.title = element_text(size=8, family='ArialMT'),
        legend.text = element_text(size=8,family='ArialMT'),
        title = element_text(size=10, family='ArialMT', face='bold'),
        panel.border = element_rect(colour='white'),
        panel.grid = element_blank()) +
  scale_fill_discrete(labels=c("Indels not in HP","Indels in HP")) + xlab("MSI sample") + 
  ylab("1bp indel counts") + ggtitle("1 bp indels in homopolymers in COAD samples with MSI")
p2
```

## More about signature contributions

Same boxplots for realtive contributions in MSI and MSS samples for all signatures simultaneously:

```{r, warning=F, fig.width=10, fig.height=3, eval=F}
mmr.type <- ifelse(rownames(mm) %in% mmr.ucsc, "MSI", "MSS")
df = data.frame(decomposition,mmr.type, sample = rownames(mm))
df2 <- melt(df,id.vars = c("mmr.type","sample"))
ggplot(data=df2, aes(x=factor(mmr.type),y=value,fill=variable)) + geom_boxplot() + facet_grid(. ~ variable,scales = "free")
```

Absolute contributions:

```{r, warning=F, echo=F, fig.width=10, fig.height=3, eval=F}
mmr.type <- ifelse(rownames(mm) %in% mmr.ucsc, "MSI", "MSS")
df = data.frame(log(t(round(new.cont.mat))),mmr.type, sample = rownames(mm))
df2 <- melt(df,id.vars = c("mmr.type","sample"))
ggplot(data=df2, aes(x=factor(mmr.type),y=value,fill=variable)) + geom_boxplot() + facet_grid(. ~ variable,scales = "free")
```

## Similarities

We calculate cosine similarities between COSMIC signatures and de novo signature set; and also between C.elegans signatures and all if the abovementioned. Cosine similarity score measures the cosine of an angle between two vectors: 

$Sim(a,b) = \frac{<a,b>}{||a||*||b||}$

```{r, eval=F}
cosine <- function(x,y) {
  x %*% y / sqrt(sum(x**2)) / sqrt(sum(y**2))
}
```

The higher the similarity, the closer the vectors are to each other.

Similarities between de novo signatures and *C. elegans* derived *mlh-1*, *pms-2* and *pole-4;pms-2* mutational patterns:

```{r, warning=F, eval=F}
tmp <- matrix(0,nrow=8,ncol=3,dimnames=list(colnames(sigs),rownames(learned.sigs)[c(2,3,5)]))
for (i in c(1:8)) 
  for (j in c(1:3)) {
    tmp[i,j] <- round(cosine(sigs[1:96,i],learned.sigs.exome[j,1:96]),2)
  }
print(tmp)
```

Similarities between de novo signatures and COSMIC cancer signatures:

```{r, warning=F, eval=F}
tmp <- matrix(0,nrow=8,ncol=30,dimnames=list(colnames(sigs),colnames(cancer_signatures)))
for (i in c(1:8)) 
  for (j in 1:30) {
    tmp[i,j] <- round(cosine(sigs[1:96,i],cancer_signatures_adj[1:96,j]),2)
  }
print(tmp)
```

## Age correlation

Check if contribution of any signature correlates with age (answer - no).

Correlation with relative contributions: 

```{r, warning=F, eval=F}
donorinfo <- rbind(read.delim("ICGC data/COAD/donor.tsv"),read.delim("ICGC data/STAD/donor.tsv"))
tmp <- vector("numeric",8)
for (i in 1:8)
  tmp[i] <- (cor(donorinfo$donor_age_at_diagnosis[match(rownames(mm)[-c( 232, 484, 494)],donorinfo$icgc_donor_id)],decomposition[-c(232,484,494),i],method='spearman'))
names(tmp) <- colnames(sigs)
print(tmp)
```

Correlation with absolute contributions: 

```{r, warning=F, eval=F}
tmp <- vector("numeric",8)
for (i in 1:8)
  tmp[i] <- (cor(donorinfo$donor_age_at_diagnosis[match(rownames(mm)[-c( 232, 484, 494)],donorinfo$icgc_donor_id)],new.cont.mat[i,-c(232,484,494)],method='spearman'))
names(tmp) <- colnames(sigs)
print(tmp)
```

## Average profiles

Plot averaged profiles of MSI samples in COAD and STAD datasets. Their cosine similarity shows that they are nearly identical.

```{r, warning=F, fig.width=10, fig.height=6, eval=F}
p1 <- plot_profiles_104(t(mm[intersect(rownames(STAD.mutation.counts),mmr.ucsc),]), boxplot = T, normalize = T, size=14)
p2 <- plot_profiles_104(t(mm[intersect(rownames(COAD.mutation.counts),mmr.ucsc),]), boxplot = T, normalize = T, size=14)
multiplot(p1,p2)
print(cosine(colMeans(mm[intersect(rownames(STAD.mutation.counts),mmr.ucsc),]),colMeans(mm[intersect(rownames(COAD.mutation.counts),mmr.ucsc),])))
```

## Cosine similarity simulations for signature comparison

Functions
```{r, eval=F}
cosine <- function(x,y) {
  return(sum(x * y) / sqrt(sum(x**2)) / sqrt(sum(y**2)))
}
cosineM <- function(X) {
  return(sapply(1:ncol(X), function(i) sapply(1:ncol(X), function(j) cosine(X[,i], X[,j]))))
}
poisI <- function(X, theta, Y){
  lambda <- as.numeric(X %*% theta)
  t(Y/lambda^2 * X) %*% X
} # Fisher information matrix for Poisson model
```

Generate uniformly distributed profiles in positive cone
```{r, eval=F}
maxs <- sample(1:5000, 1000, replace=T)
X <- sapply(1:1000, function(x) runif(n = 104, min = 0, max=maxs[x]))
X <- apply(X,2,function(y) y/sum(y))
XM <- cosineM(X)
df = data.frame(x = 1:length(XM[upper.tri(XM)]), val = XM[upper.tri(XM)])
p <- ggplot(data = df, aes(val)) + 
  geom_histogram(binwidth = 0.005, col='black', fill='white') + 
  theme_bw() + theme(panel.grid = element_blank(), text = element_text(family='ArialMT'), panel.border = element_rect(colour='white')) +
  geom_vline(xintercept = quantile(as.vector(XM[upper.tri(XM)]), 0.95), col='red', linetype='dashed') +
  ggtitle('Distribution of similarities between uniform random vectors from positive cone') +
  xlab('Cosine similarity score') + ylab('Frequency') + xlim(c(0,1))
ggsave(plot=p, height=4,width=7, filename = '~/Documents/uniform_sim.pdf', useDingbats=FALSE)
```

## Distribution of angles in COSMIC signatures

Get the histogram and heatmap
```{r, eval=F}
cosmic <- cosineM(cancer_signatures)
hist(cosmic[upper.tri(cosmic)], prob=T, breaks=20, main='Similarities between COSMIC signatures', xlab = 'Cosine similarity')
lines(density(cosmic[upper.tri(cosmic)], adjust=2), lty="dotted") 

# Heatmap
image.plot(cosmic)
for (x in 1:ncol(cosmic))
  for (y in 1:ncol(cosmic))
    text((x-1)/(ncol(cosmic)-1), (y-1)/(ncol(cosmic)-1), sprintf("%0.2f", cosmic[x,y]))
# ggplot histogram
df = data.frame(x = 1:length(cosmic[upper.tri(cosmic)]), val=cosmic[upper.tri(cosmic)])
p <- ggplot(data = df, aes(val)) + 
  geom_histogram(binwidth = 0.05, col='black', fill='white') + 
  theme_bw() + theme(panel.grid = element_blank(), text = element_text(family='ArialMT'), panel.border = element_rect(colour='white')) +
  geom_vline(xintercept = 0.8, col='red', linetype='dashed') +
  ggtitle('Distribution of similarities between COSMIC cancer signatures') +
  xlab('Cosine similarity score') + ylab('Frequency') + xlim(c(0,1))
ggsave(plot=p, height=4,width=7, filename = '~/Documents/cosmic_sim.pdf', useDingbats=FALSE)
```

Another approach: simulation of signatures using Dirichlet(1) distribution, it generates signatures uniformly in simplex => cosines are centered in 0.5, the higher the dimension - the higher the peak in 0.5. Already at 96 the distribution is close to normal.
```{r, eval=F}
library(bayesm)
# 96D case - already roughly normal, but centre not exactly in 0.5 (0.51)
Dr1 <- sapply(1:1000, function(x) rdirichlet(alpha = rep(1,96)))
Dr1M <- cosineM(Dr1)
hist(Dr1M[upper.tri(Dr1M)],breaks=100, main='Distribution of similarities for random signatures', xlab='Similarity score')
abline(v = quantile(as.vector(Dr1M[upper.tri(Dr1M)]), 0.95), col='red', lty=2) # (0.61)
# 1000D case - normal, centre very close to 0.5
Dr1000 <- sapply(1:1000, function(x) rdirichlet(alpha = rep(1,1000)))
Dr1000M <- cosineM(Dr1000)
hist(Dr1000M[upper.tri(Dr1000M)],breaks=100, main='Distribution of similarities for random signatures', xlab='Similarity score')
abline(v = quantile(as.vector(Dr1000M[upper.tri(Dr1000M)]), 0.95), col='red', lty=2)
```

Now draw a signature from bootstrapped COADSTAD signatures and compare to randomly drawn within CI worm signature
```{r, eval=F}
load('~/profiles_and_decomposition.RData')
mm <- rbind(COAD.mutation.counts,STAD.mutation.counts)
mut_mat = t(mm) + 0.0001
res <- nmf(mut_mat,rank=8,seed=123456,method='brunet')
defsigs <- NMF::basis(res)
defsigs <- defsigs[,c(8,1,2,5,4,7,3,6)]
colnames(defsigs) <- c("Clock-1", "Clock-2", "POLE", "17-like", "MMR-1", "MMR-2", "MMR-3", "SNP")
```

Bootstrapping
```{r, eval=F}
sigs = list()
for (i in 1:505) {
  mm = rbind(COAD.mutation.counts,STAD.mutation.counts)
  mut_mat <- t(mm[-i,]) + 0.0001
  res <- nmf(mut_mat, rank = 8, method='brunet')
  sigs[[i]] <- NMF::basis(res)
}
```

Check out stability of signatures in those bootstraps
```{r, eval=F}
similarity_hist <- function(defsigs,set,k) {
  hist(sapply(1:ncol(set[[k]]), function(s) cosine(defsigs[,k],set[[k]][,s])), breaks=100)
}
jkset = list()
for (k in 1:8) {
  jkset[[k]] <- sapply(1:505, function(i) 
    sigs[[i]][,which.max(sapply(1:8, function(j) cosine(sigs[[i]][,j],defsigs[,k])))] / sum(sigs[[i]][,which.max(sapply(1:8, function(j) cosine(sigs[[i]][,j],defsigs[,k])))]))
}
similarity_hist(defsigs,jkset,5)
summary(sapply(1:ncol(jkset[[5]]), function(s) cosine(defsigs[,5],jkset[[5]][,s])))
```

Now take the worm sigs randomly from their CIs
```{r, eval=F}
load('~/Learned_signatures_indel.RData')
learned.sigs <- nmSolve(worm.donor.mut.mat[,1:96],small.X,maxIter=10000, tol = 1e-06, div.err = 1e-10)
Y = worm.donor.mut.mat
mut_matrix_lower <- matrix(NA,nrow=96,ncol=nrow(learned.sigs),dimnames=list(colnames(learned.sigs)[1:96],row.names(learned.sigs)))
mut_matrix_upper <- matrix(NA,nrow=96,ncol=nrow(learned.sigs),dimnames=list(colnames(learned.sigs)[1:96],row.names(learned.sigs)))
library(MASS)
for (i in 1:96) {
  cov.mat <- poisI(as.matrix(small.X),learned.sigs[,i],Y[,i])
  to.keep <- colnames(cov.mat)[which(diag(cov.mat)!=0)]
  if (length(to.keep)!=0) {
    cov.mat <- cov.mat[to.keep,to.keep]
    SVD <- svd(cov.mat)$d
    if (length(which(SVD<1e-6))>0)
      cov.mat <- ginv(cov.mat)
    if (length(which(SVD<1e-6))==0)
      cov.mat <- solve(cov.mat)
    row.names(cov.mat) <- to.keep
    colnames(cov.mat) <- to.keep
    sterr <- sqrt(diag(cov.mat))
    mut_matrix_lower[i,to.keep] <- learned.sigs[to.keep,i] - sterr * qt(0.975,30)
    mut_matrix_upper[i,to.keep] <- learned.sigs[to.keep,i] + sterr * qt(0.975,30)
    mut_matrix_lower[i,to.keep][mut_matrix_lower[i,to.keep]<0] <- 0
  }
} 
```

Adjust for trinucleotide difference
```{r, eval=F}
load('exome.RData')
learned.sigs.exome <- learned.sigs[,1:96]
mut_matrix_lower.exome <- t(mut_matrix_lower)
mut_matrix_upper.exome <- t(mut_matrix_upper)
for (i in 1:nrow(learned.sigs.exome)) {
  learned.sigs.exome[i,] <- learned.sigs.exome[i,] / trinucleotide.freq.factor.ex[types]
  mut_matrix_lower.exome[i,1:96] <- mut_matrix_lower.exome[i,1:96] / trinucleotide.freq.factor.ex[types] # + learned.sigs.exome[i,1:96] - learned.sigs[i,1:96]
  mut_matrix_upper.exome[i,1:96] <- mut_matrix_upper.exome[i,1:96] / trinucleotide.freq.factor.ex[types] # + learned.sigs.exome[i,1:96] - learned.sigs[i,1:96]
}
mut_matrix_lower.exome[mut_matrix_lower.exome<0] <- 0
# plot signatures with CIs
plot_96_profile_CI(mut_matrix = t(learned.sigs[c(2:3,5),1:96]), mut_matrix_lower = mut_matrix_lower[,c(2:3,5)], mut_matrix_upper = mut_matrix_upper[,c(2:3,5)], CI=TRUE)
plot_96_profile_CI(mut_matrix = t(learned.sigs.exome[c(2:3,5),1:96]), mut_matrix_lower = t(mut_matrix_lower.exome[c(2:3,5),]), mut_matrix_upper = t(mut_matrix_upper.exome[c(2:3,5),]), CI=TRUE)
```

Now draw worm and signatures randomly from their confidence intervals.
Similarities for fixed signatures:
```{r, eval=F}
cosine(learned.sigs.exome[3,1:96],defsigs[1:96,5]) # # 0.85
cosine(learned.sigs.exome[2,1:96],defsigs[1:96,5]) # # 0.81
cosine(learned.sigs.exome[3,-c(35,39,43,47,97:104)],defsigs[-c(35,39,43,47,97:104),5]) # 0.92
cosine(learned.sigs.exome[2,-c(35,39,43,47,97:104)],defsigs[-c(35,39,43,47,97:104),5]) # 0.90
```
Drawing random signatures:
```{r, eval=F}
random_signature <- function(sig, sig_low, sig_up) {
  res <- sapply(1:length(sig), function(i) rnorm(1,mean=sig[i],sd=(sig_up[i]-sig[i])/2))
  res[res<0] <- 0
  res[is.nan(res)] <- sig[is.nan(res)]
  return(res)
}
```

**pole4;pms-2** double mutant versus single signatures:
```{r, eval=F}
a1 <- sapply(1:1000, function(s) cosine(random_signature(sig=learned.sigs[2,1:96],sig_low = mut_matrix_lower[,2],sig_up = mut_matrix_upper[,2]),
                                        random_signature(sig=learned.sigs[5,1:96],sig_low = mut_matrix_lower[,5],sig_up = mut_matrix_upper[,5])))
a2 <- sapply(1:1000, function(s) cosine(random_signature(sig=learned.sigs[3,1:96],sig_low = mut_matrix_lower[,3],sig_up = mut_matrix_upper[,3]),
                                        random_signature(sig=learned.sigs[5,1:96],sig_low = mut_matrix_lower[,5],sig_up = mut_matrix_upper[,5])))
a3 <- sapply(1:1000, function(s) cosine(random_signature(sig=learned.sigs[2,1:96],sig_low = mut_matrix_lower[1:96,2],sig_up = mut_matrix_upper[1:96,2]),
                                        random_signature(sig=learned.sigs[3,1:96],sig_low = mut_matrix_lower[1:96,3],sig_up = mut_matrix_upper[1:96,3])))

df = data.frame(x = 1:length(a1), a1, a2, a3)
df = melt(df, id.vars = "x")
levels(df$variable) <- c('mlh-1 vs pole-4;pms-2', 'pms-2 vs pole-4;pms-2', 'mlh-1 vs pms-2')
df2 <- data.frame(cos = c(cosine(learned.sigs[2,1:96],learned.sigs[5,1:96]),
                           cosine(learned.sigs[3,1:96],learned.sigs[5,1:96]),
                           cosine(learned.sigs[2,1:96],learned.sigs[3,1:96])),
                  variable = c('mlh-1 vs pole-4;pms-2', 'pms-2 vs pole-4;pms-2', 'mlh-1 vs pms-2'))
p <- ggplot(data = df, aes(value)) + 
  geom_histogram(binwidth = 0.01, col='black', fill='white') + 
  facet_grid(variable ~ ., scales='fixed') +
  theme_bw() + theme(panel.grid = element_blank(), text = element_text(family='ArialMT'), strip.background = element_rect(colour='white', fill='white'),
                     panel.border = element_rect(colour = 'white'), strip.text = element_text(size=12)) +
  geom_vline(data=df2, aes(xintercept=cos), col='red', linetype='dashed') +
  ggtitle('Distribution of similarities between C. elegans signatures') +
  xlab('Cosine similarity score') + ylab('Frequency') + xlim(c(0,1))
p
```

**mlh-1** and **pms-2** vs MMR-1
```{r, eval=F}
a1 <- sapply(1:505, function(s) cosine(random_signature(sig=learned.sigs.exome[3,1:96],sig_low = mut_matrix_lower.exome[3,],sig_up = mut_matrix_upper.exome[3,]),jkset[[5]][-c(97:104),s]))
a2 <- sapply(1:505, function(s) cosine(random_signature(sig=learned.sigs.exome[3,c(1:96)[-c(35,39,43,47)]],sig_low = mut_matrix_lower.exome[3,-c(35,39,43,47)],
                                                                   sig_up = mut_matrix_upper.exome[3,-c(35,39,43,47)]),jkset[[5]][-c(35,39,43,47,97:104),s]))
b1 <- sapply(1:505, function(s) cosine(random_signature(sig=learned.sigs.exome[2,1:96],sig_low = mut_matrix_lower.exome[3,],
                                                                   sig_up = mut_matrix_upper.exome[3,]),jkset[[5]][-c(97:104),s]))
b2 <- sapply(1:505, function(s) cosine(random_signature(sig=learned.sigs.exome[2,c(1:96)[-c(35,39,43,47)]],sig_low = mut_matrix_lower.exome[3,-c(35,39,43,47)],
                                                                   sig_up = mut_matrix_upper.exome[3,-c(35,39,43,47)]),jkset[[5]][-c(35,39,43,47,97:104),s]))


a1 <- sapply(1:505, function(s) cosine(learned.sigs.exome[3,1:96],jkset[[5]][-c(97:104),s]))
a2 <- sapply(1:505, function(s) cosine(learned.sigs.exome[3,c(1:96)[-c(35,39,43,47)]],jkset[[5]][-c(35,39,43,47,97:104),s]))
b1 <- sapply(1:505, function(s) cosine(learned.sigs.exome[2,1:96],jkset[[5]][-c(97:104),s]))
b2 <- sapply(1:505, function(s) cosine(learned.sigs.exome[2,c(1:96)[-c(35,39,43,47)]],jkset[[5]][-c(35,39,43,47,97:104),s]))


df = data.frame(x = 1:length(a1), a1, a2)
df = melt(df, id.vars = "x")
levels(df$variable) <- c('Full signatures', 'Without C>T at CpG sites')
df2 <- data.frame(cos = c(cosine(learned.sigs.exome[3,1:96],defsigs[1:96,5]),
                          cosine(learned.sigs.exome[3,-c(35,39,43,47,97:104)],defsigs[-c(35,39,43,47,97:104),5])),
                  variable =c('Full signatures', 'Without C>T at CpG sites'))
p <- ggplot(data = df, aes(value)) + 
  geom_histogram(binwidth = 0.01, col='black', fill='white') + 
  facet_grid(. ~ variable, scales='fixed') +
  theme_bw() + theme(panel.grid = element_blank(), text = element_text(family='ArialMT'), strip.background = element_rect(colour='white', fill='white'),
                     panel.border = element_rect(colour = 'white'), strip.text = element_text(size=12)) +
  geom_vline(data=df2, aes(xintercept = cos), col='red', linetype='dashed') +
  ggtitle('Distribution of similarities between pms-2 and MMR-1 signatures') +
  xlab('Cosine similarity score') + ylab('Frequency') + xlim(c(0,1))
p

df = data.frame(x = 1:length(a1), b1, b2)
df = melt(df, id.vars = "x")
levels(df$variable) <- c('Full signatures', 'Without C>T at CpG sites')
df2 <- data.frame(cos = c(cosine(learned.sigs.exome[2,1:96],defsigs[1:96,5]),
                          cosine(learned.sigs.exome[2,-c(35,39,43,47,97:104)],defsigs[-c(35,39,43,47,97:104),5])),
                  variable =c('Full signatures', 'Without C>T at CpG sites'))
p <- ggplot(data = df, aes(value)) + 
  geom_histogram(binwidth = 0.01, col='black', fill='white') + 
  facet_grid(. ~ variable, scales='fixed') +
  theme_bw() + theme(panel.grid = element_blank(), text = element_text(family='ArialMT'), strip.background = element_rect(colour='white', fill='white'),
                     panel.border = element_rect(colour = 'white'), strip.text = element_text(size=12)) +
  geom_vline(data=df2, aes(xintercept = cos), col='red', linetype='dashed') +
  ggtitle('Distribution of similarities between mlh-1 and MMR-1 signatures') +
  xlab('Cosine similarity score') + ylab('Frequency') + xlim(c(0,1))
p
```

# SNP contamination signature
Checking the presence of human population SNPs identified in 1000 genome project in COAD/STAD samples providing evidence for de novo signature 8 representing SNP contamination.

Get the SNPs
```{r}
library('SNPlocs.Hsapiens.dbSNP144.GRCh37')
snps <- SNPlocs.Hsapiens.dbSNP144.GRCh37
```

Find substitutions in COAD variants
```{r, eval=F}
regs <- sapply(vcf_list_COAD, function(vcf) {
  vcf[width(vcf$REF)==1 & width(unlist(vcf$ALT))==1,]
})
# Adjust chromosome names
for (sample in names(regs)) {
  seqlevels(regs[[sample]]) <- substr(seqlevels(regs[[sample]]),4,nchar(seqlevels(regs[[sample]])))
}
# Intersect SNPs and substitutions
snps_in_regs <- list()
for (i in seq_along(vcf_list_COAD)) {
  genome(regs[[i]]) <- "GRCh37.p13"
  snps_in_regs[[i]] <- snpsByOverlaps(snps,regs[[i]],minoverlap=1)
  inds <- match(pos(snps_in_regs[[i]]),start(regs[[i]]))
  alts <- as.character(unlist(regs[[i]][inds]$ALT))
  snpalts <- as.character(IUPAC_CODE_MAP[snps_in_regs[[i]]$alleles_as_ambig])
  if (length(alts)==0) next
  snps_in_regs[[i]] <- snps_in_regs[[i]][sapply(1:length(inds), function(j) {
    grepl(alts[j],snpalts[j])
  })]
}
names(snps_in_regs) <- names(regs)
```

Find substitutions in STAD variants

```{r, eval=F}
STregs <- sapply(vcf_list_STAD, function(vcf) {
  vcf[width(vcf$REF)==1 & width(unlist(vcf$ALT))==1,]
})
# Intersect SNPs and substitutions
STsnps_in_regs <- list()
for (i in seq_along(vcf_list_STAD)) {
  seqlevels(STregs[[i]]) <- substr(seqlevels(STregs[[i]]),4,nchar(seqlevels(STregs[[i]])))
  genome(STregs[[i]]) <-  "GRCh37.p13"
  STsnps_in_regs[[i]] <- snpsByOverlaps(snps,STregs[[i]],minoverlap=1)
  print(i)
  inds <- match(pos(STsnps_in_regs[[i]]),start(STregs[[i]]))
  alts <- as.character(unlist(STregs[[i]][inds]$ALT))
  snpalts <- as.character(IUPAC_CODE_MAP[STsnps_in_regs[[i]]$alleles_as_ambig])
  if (length(alts)==0) next
  STsnps_in_regs[[i]] <- STsnps_in_regs[[i]][sapply(1:length(inds), function(j) {
    grepl(alts[j],snpalts[j])
  })]
}
names(STsnps_in_regs) <- names(STregs)
```

Plot the fraction of potential SNPs to all variants per sample
```{r, eval=F}
hist( sapply(STsnps_in_regs,length) / sapply(STregs,length), breaks=20)
```

Get coding and non-coding SNPs from 1000 Genomes project (http://www.internationalgenome.org/category/vcf/)
```{r, eval=F}
coding_vcf <- readVcf("ALL.wgs.integrated_phase1_release_v3_coding_annotation.20101123.snps_indels.sites.vcf")
non_coding_vcf <- readVcf("ALL.wgs.integrated_phase1_release_v3_noncoding_annotation_20120330.20101123.snps_indels_sv.sites.vcf.gz")
```

Count the coding SNPs
```{r, eval=F}
regs <- c(regs, STregs)
cod_snps_by_overlaps <- sapply(regs, function(reg) {
  tmp <- findOverlaps(granges(coding_vcf), reg)
  alts <- as.character(unlist(reg[subjectHits(tmp)]$ALT))
  snpalts <- as.character(unlist(granges(coding_vcf)$ALT))[queryHits(tmp)]
  tmp[alts==snpalts]
})
#non_cod_snps_by_overlaps <- sapply(regs, function(reg) findOverlaps(non_coding_vcf, reg))
counts <- sapply(cod_snps_by_overlaps, function(x) 
  return(c(length(grep(":synonymous:",unlist(info(coding_vcf[queryHits(x)])$VA))),
           length(grep(":nonsynonymous:",unlist(info(coding_vcf[queryHits(x)])$VA))))))
```

Check nonsynonymous to synonimous
```{r, eval=F}
length(grep(":nonsynonymous:",unlist(info(coding_vcf)$VA))) / length(grep(":synonymous:",unlist(info(coding_vcf)$VA)))
```

Take the ones with signature 8
```{r, eval=F}
suspects <- rownames(decomposition)[decomposition[,8]>0.3]
barplot(counts[2,suspects[counts[1,suspects]!=0]] / (counts[1,suspects[counts[1,suspects]!=0]]))
```

# Homopolymer analysis in **C. elegans**

Define homopolymers:

```{r, eval=F}
motif_6bp_HP <- DNAStringSet(c("BAAAAAAB", "VTTTTTTV", "HGGGGGGH", "DCCCCCCD"))
homopolymer_pool_length_4to35 <- lapply(seq(4,35),function(y) lapply(c("A","C","G","T"),function(x)
  if (x=="A") {paste0("B",paste(rep(x,y),collapse=""),"B")
  }
  else
    if (x=="T") {paste0("V",paste(rep(x,y),collapse=""),"V")
    }
  else
    if (x=="G") {paste0("H",paste(rep(x,y),collapse=""),"H")
    }
  else
    if (x=="C") {paste0("D",paste(rep(x,y),collapse=""),"D")
    }
))
whomopolymers <- unlist(homopolymer_pool_length_4to35)
list(whomopolymers)
# turn into DNA stringset #
pattern_all <- DNAStringSet(whomopolymers)
```

Find homopolymers in each chromosome:

```{r, eval=F}
whits_all <- lapply(WBcel235, function(chr) sapply(pattern_all, matchPattern, subject=chr, fixed=FALSE))
wlengths <- lapply(whits_all, function(chr) sapply(chr, length)) # numbers of homopolymers on each chromosome
wnonzero_hits_all <- lapply(1:7, function(i) whits_all[[i]][which(wlengths[[i]]>0)]) # 89 82 85 87 13 92 89 - numbers of classes for each chromosome
wnonzero_lengths <- lapply(wnonzero_hits_all, function(chr) sapply(chr, length)) # numbers of homopolymers on each chromosome for non empty classes only
names(wnonzero_hits_all) <- c("I","II","III","IV","MtDNA","V","X")
```

Generate GRanges object out of list of all motif hits:

```{r, eval=F}
sites.gr.worm <- lapply(1:7, function(i) do.call("c",lapply(lapply(wnonzero_hits_all[[i]], as, "IRanges"),
                                                            GRanges,seqnames=names(wnonzero_hits_all)[i])))
```

Add corresponding motif sequences and genomic homopolymer length to the GRanges object:

```{r, eval=F}
for (j in 1:7) {
  sites.gr.worm[[j]]$pattern.searched <- rep(as.character(pattern_all), sapply(whits_all[[j]], length))
  sites.gr.worm[[j]]$motif.found <- unlist(lapply(wnonzero_hits_all[[j]], as.character))
  sites.gr.worm[[j]]$pattern.length <- unlist(lapply(wnonzero_hits_all[[j]], width)) # includes the two flanking bases
  sites.gr.worm[[j]]$homopolymer.length <- (sites.gr.worm[[j]]$pattern.length)-2 # remove 2 bases, 5' and 3' are not part of homopolymer
}
```

Combine information from all Chromosomes into one GRanges object showing all hits across the genome:

```{r, eval=F}
sites.gr_all.worm <- do.call("c",sites.gr.worm[-5]) # exclude MtDNA
genome(sites.gr_all.worm) <- "WBcel235" # add genome info
```

Create a dataframe with homopolymer information:

```{r, eval=F}
homopolymers.worm <- as.data.frame(table(all_sites_worm$pattern.searched))
homopolymers.worm$length <- width(as.character(homopolymers.worm$Var1))-2 # include length of motif, -2 removes flanking 5'and 3' bases
homopolymers.worm$base <- ifelse(grepl("AAA",homopolymers.worm$Var1),"A",ifelse(grepl("TTT", homopolymers.worm$Var1), "T", ifelse(grepl("GGG", homopolymers.worm$Var1), "G", "C")))
homopolymer_frequency_by_length_worm <- homopolymers.worm %>%
  group_by(length) %>%
  summarise(total_frequency =sum(Freq))
homopolymer_frequency_by_length_worm$total_number_of_bases <- (homopolymer_frequency_by_length_worm$length)*as.numeric(homopolymer_frequency_by_length_worm$total_frequency) # calculate the number of bases for each HP
```

C. elegans indels in homopolymers: ipload vcf from C. elegans with indels in a list of vcfs called MMRvcf.

```{r, eval=F}
# Select 1-bp indels
indel1bp <- sapply(MMRvcf, function(vcf) vcf[abs(width(granges(vcf)$REF)-width(unlist(granges(vcf)$ALT)))==1 &
                                               (width(granges(vcf)$REF)==1 | width(unlist(granges(vcf)$ALT))==1)])
barplot(sapply(indel1bp, length))
```
```{r,eval=F}
# Prepare the Ranges files
indel1bp <- sapply(indel1bp, rowRanges)
for (k in 1:length(indel1bp)) {
  indel1bp[[k]]$insertion <- width(unlist(indel1bp[[k]]$ALT))-1
  indel1bp[[k]]$deletion <- width(indel1bp[[k]]$REF)-1
}
```

Now check the intersections of homopolymers and indels; watch out for cases when an indel is between 2 homopolymers and can be counted twice.

```{r}
worm_hit_list <- sapply(indel1bp, function(x) findOverlaps(x,sites.gr_all.worm))
indels.to.check.list <- sapply(worm_hit_list, function(x) unique(as.matrix(x)[,1])[which(table(as.matrix(x)[,1])>1)])
for (t in which(sapply(indels.to.check.list,length)>0)) {
  to.delete.final <- NULL
  indels.to.check <- indels.to.check.list[[t]]
  for (j in 1:length(indels.to.check)) {
    homopolymers.to.check <- as.matrix(worm_hit_list[[t]])[which(as.matrix(worm_hit_list[[t]])[,1] == (indels.to.check[j])),2]
    
    homopolymers.types <- sapply(sites.gr_all.worm[homopolymers.to.check]$pattern.searched,function(x) substr(x,2,2))
    
    ref <- unlist(strsplit(as.data.frame(indel1bp[[t]][indels.to.check[j]])$REF,split = ""))
    alt <- unlist(strsplit(as.character(as.data.frame(indel1bp[[t]][indels.to.check[j]])$ALT[[1]]),split = ""))
    
    if (indel1bp[[t]][indels.to.check[j]]$insertion==1) {
      indel.type <- setdiff(alt,ref)
      to.delete <- which(as.matrix(worm_hit_list[[t]])[,1]==indels.to.check[j]) [homopolymers.types != indel.type]
      if (length(to.delete)==0) {
        if (indel.type==alt[1]) to.delete <- which(as.matrix(worm_hit_list[[t]])[,1]==indels.to.check[j])[2]
        if (indel.type==alt[2]) to.delete <- which(as.matrix(worm_hit_list[[t]])[,1]==indels.to.check[j])[1]
      }
    }
    if (indel1bp[[t]][indels.to.check[j]]$deletion==1) {
      indel.type <- setdiff(ref,alt)
      to.delete <- which(as.matrix(worm_hit_list[[t]])[,1]==indels.to.check[j]) [homopolymers.types != indel.type]
      if (length(to.delete)==0) {
        if (indel.type==ref[1]) to.delete <- which(as.matrix(worm_hit_list[[t]])[,1]==indels.to.check[j])[2]
        if (indel.type==ref[2]) to.delete <- which(as.matrix(worm_hit_list[[t]])[,1]==indels.to.check[j])[1]
      }
    }
    to.delete.final <- c(to.delete.final,to.delete)
  }
  if (length(to.delete.final)>0) worm_hit_list[[t]] <- worm_hit_list[[t]][-to.delete.final]
}
worm_hit_list <- worm_hit_list[sapply(worm_hit_list,length)>0]
```

Turn the hist into a list of dataframes:

```{r, eval=F}
worm_hits_data_frame <- list()
for (i in 1:length(worm_hit_list)) {
  queries <- unique(queryHits(worm_hit_list[[i]]))
  idx <- sapply(queries, function(x) which(queryHits(worm_hit_list[[i]])==x)[1])
  idxs <- subjectHits(worm_hit_list[[i]])[idx]
  motifs <- DataFrame(sites.gr_all.worm$motif.found[idxs])
  length <- unlist(lapply(sites.gr_all.worm$motif.found[idxs], width))-2
  worm_hits_data_frame[[i]] <- as.data.frame(indel1bp[[names(worm_hit_list)[i]]][queries,])
  worm_hits_data_frame[[i]]$motif.found <- sites.gr_all.worm$motif.found[idxs]
  worm_hits_data_frame[[i]]$motif.length <- width(worm_hits_data_frame[[i]]$motif.found)
  worm_hits_data_frame[[i]]$homopolymer.length <- worm_hits_data_frame[[i]]$motif.length-2
  print(i)
}
names(worm_hits_data_frame) <- names(worm_hit_list)
```

Add homopolymer type:

```{r, eval=F}
for (i in 1:length(worm_hits_data_frame)) {
  worm_hits_data_frame[[i]]$base <- ifelse(grepl("AAA",worm_hits_data_frame[[i]]$motif.found),"A",
                                           ifelse(grepl("TTT", worm_hits_data_frame[[i]]$motif.found), "T", 
                                                  ifelse(grepl("GGG", worm_hits_data_frame[[i]]$motif.found), "G", "C")))
  worm_hits_data_frame[[i]]$hpn <- paste0(worm_hits_data_frame[[i]]$homopolymer.length,worm_hits_data_frame[[i]]$base)
}
```

Group the data:

```{r, eval=F}
mlh1_20 = do.call("rbind",worm_hits_data_frame[c("CD0134a","CD0134c","CD0134d")])
pms2_20 = do.call("rbind",worm_hits_data_frame[c("CD0135a","CD0135c","CD0135d")])
pms2_10 = do.call("rbind",worm_hits_data_frame[c("CD0244a","CD0244c","CD0244d")])
pole4pms2_10 = do.call("rbind",worm_hits_data_frame[c("CD0246d","CD0246e")])
# Variant calling in long homopymers is infeasible (too few reads covering the whole homopolymer)
mlh1_20 = mlh1_20[mlh1_20$homopolymer.length<19,]
pms2_20 = pms2_20[pms2_20$homopolymer.length<19,]
pms2_10 = pms2_10[pms2_10$homopolymer.length<19,]
pole4pms2_10 = pole4pms2_10[pole4pms2_10$homopolymer.length<19,]
```

Plot raw frequencies and ratios:

```{r, eval=F}
total_frequency = homopolymer_frequency_by_length_worm$total_frequency[1:15]
to.show <- data.frame(homopolymer.length = 4:(length(total_frequency)+3), 
                      pms2.20 = vector("numeric",length(total_frequency)),
                      mlh1.20 = vector("numeric",length(total_frequency)),
                      pms2.10 = vector("numeric",length(total_frequency)),
                      pole4pms2.10 = vector("numeric",length(total_frequency)))
to.show$mlh1.20[as.numeric(names(table(mlh1_20$homopolymer.length)))-3] <- (table(mlh1_20$homopolymer.length)/3)
to.show$pms2.20[as.numeric(names(table(pms2_20$homopolymer.length)))-3] <- (table(pms2_20$homopolymer.length)/3) 
to.show$pms2.10[as.numeric(names(table(pms2_10$homopolymer.length)))-3] <- (table(pms2_10$homopolymer.length)/3)
to.show$pole4pms2.10[as.numeric(names(table(pole4pms2_10$homopolymer.length)))-3] <- round(table(pole4pms2_10$homopolymer.length)/2)
```

Plot absolute numbers:

```{r, eval=F}
df <- melt(to.show,id.vars = "homopolymer.length")
ggplot(data=df,aes(x=homopolymer.length,y=value)) + geom_bar(stat='identity', width = 0.75) +
  facet_grid(. ~ factor(variable,levels=c("mlh1.20","pms2.20","pms2.10","pole4pms2.10"))) +theme_bw() +
  ylab("Number of 1 bp indels") + xlab("Homopolymer length") + 
  theme(axis.text = element_text(size=12, family='ArialMT'), 
        axis.title = element_text(size=16, family='ArialMT'), 
        strip.text = element_text(size=16, family='ArialMT'),
        panel.grid = element_blank(), strip.background = element_blank())
```

Plot relative frequencies:

```{r, eval=F}
for (i in 2:ncol(to.show)) to.show[,i] = to.show[,i] / total_frequency
df <- melt(to.show,id.vars = "homopolymer.length")
ggplot(data=df,aes(x=homopolymer.length,y=value)) + 
  facet_grid(. ~ factor(variable,levels=c("mlh1.20","pms2.20","pms2.10","pole4pms2.10"))) +
  geom_point() + geom_line() + theme_bw() + 
  ylab("Ratio \n Mutations/HP") + xlab("Homopolymer length") +
  theme(axis.text = element_text(size=12), axis.title = element_text(size=16), strip.text = element_text(size=16))
```

Apply generalized additive models with splines and plot the results:

```{r, eval=F}
total_frequency = homopolymer_frequency_by_length_worm$total_frequency[1:15]
to.show <- data.frame(homopolymer.length = 4:(length(total_frequency)+3), 
                      pms2 = vector("numeric",length(total_frequency)),
                      mlh1= vector("numeric",length(total_frequency)),
                      pole4pms2= vector("numeric",length(total_frequency)),
                      pms210 = vector("numeric",length(total_frequency)),
                      total_frequency = total_frequency,
                      total_number_of_bases = homopolymer_frequency_by_length_worm$total_number_of_bases[1:15])
to.show$mlh1[as.numeric(names(table(mlh1_20$homopolymer.length)))-3] <- round(table(mlh1_20$homopolymer.length)/3)
to.show$pms2[as.numeric(names(table(pms2_20$homopolymer.length)))-3] <- round(table(pms2_20$homopolymer.length)/3)
to.show$pms210[as.numeric(names(table(pms2_10$homopolymer.length)))-3] <- round(table(pms2_10$homopolymer.length)/3)
to.show$pole4pms2[as.numeric(names(table(pole4pms2_10$homopolymer.length)))-3] <- round(table(pole4pms2_10$homopolymer.length)/2)
```

**mlh-1** mutants, generation 20
```{r, eval=F}
fit <- gam(cbind(mlh1, total_frequency-mlh1) ~ s(homopolymer.length, 6), data=to.show, family=binomial)
plot(to.show$homopolymer.length, to.show$mlh1/to.show$total_frequency, xlim=c(4,18), ylim=c(0,0.015), ylab="Ratio \n Mutations/HP", pch=16, col='grey', main='mlh-1',font.main=3, xlab="Homopolymer length")
segments(to.show$homopolymer.length, qbeta(0.025, .5+to.show$mlh1, .5+to.show$total_frequency- to.show$mlh1),
         to.show$homopolymer.length, qbeta(0.975, .5+to.show$mlh1, .5+to.show$total_frequency- to.show$mlh1), col='grey')
p <-  predict(fit, type='link', se.fit=TRUE)
lines(to.show$homopolymer.length, exp(p$fit), col='red', lwd=2)
lines(to.show$homopolymer.length, exp( p$fit +  2*p$se.fit), col='red', lty=2)
lines(to.show$homopolymer.length, exp(p$fit -  2* p$se.fit), col='red', lty=2)
lines(to.show$homopolymer.length, exp(p$fit), col='#FF00000d', lwd=2)
for(q in seq(0.025, 0.475,0.025))
  polygon(c(to.show$homopolymer.length, rev(to.show$homopolymer.length)), c(exp(qnorm(q, p$fit, p$se.fit)), exp(rev(qnorm(1-q, p$fit, p$se.fit)))), border=NA, col='#FF00000d')
```

**pms-2** mutants, generation 20
```{r, eval=F}
fit <- gam(cbind(pms2, total_frequency-pms2) ~ s(homopolymer.length, 6), data=to.show, family=binomial)
plot(to.show$homopolymer.length, to.show$pms2/to.show$total_frequency, yaxt="n",
     xlim=c(4,18), ylim=c(0,0.015), ylab=NA,
     pch=16, col='grey', main='pms-2',font.main=3, xlab="Homopolymer length")
segments(to.show$homopolymer.length, qbeta(0.025, .5+to.show$pms2, .5+to.show$total_frequency- to.show$pms2),
         to.show$homopolymer.length, qbeta(0.975, .5+to.show$pms2, .5+to.show$total_frequency- to.show$pms2), col='grey')
p <-  predict(fit, type='link', se.fit=TRUE)
lines(to.show$homopolymer.length, exp(p$fit), col='red', lwd=2)
lines(to.show$homopolymer.length, exp( p$fit +  2*p$se.fit), col='red', lty=2)
lines(to.show$homopolymer.length, exp(p$fit -  2* p$se.fit), col='red', lty=2)
lines(to.show$homopolymer.length, exp(p$fit), col='#FF00000d', lwd=2)
for(q in seq(0.025, 0.475,0.025))
  polygon(c(to.show$homopolymer.length, rev(to.show$homopolymer.length)), c(exp(qnorm(q, p$fit, p$se.fit)), exp(rev(qnorm(1-q, p$fit, p$se.fit)))), border=NA, col='#FF00000d')
```

**pms-2** mutants, generation 10
```{r, eval=F}
fit <- gam(cbind(pms210, total_frequency-pms210) ~ s(homopolymer.length, 6), data=to.show, family=binomial)
plot(to.show$homopolymer.length, to.show$pms210/to.show$total_frequency, yaxt="n", xlim=c(4,18), ylim=c(0,0.015), ylab=NA, pch=16, col='grey', main='pms-2 10',font.main=3, xlab="Homopolymer length")
segments(to.show$homopolymer.length, qbeta(0.025, .5+to.show$pms210, .5+to.show$total_frequency- to.show$pms210),
         to.show$homopolymer.length, qbeta(0.975, .5+to.show$pms210, .5+to.show$total_frequency- to.show$pms210), col='grey')
p <-  predict(fit, type='link', se.fit=TRUE)
lines(to.show$homopolymer.length, exp(p$fit), col='red', lwd=2)
lines(to.show$homopolymer.length, exp( p$fit +  2*p$se.fit), col='red', lty=2)
lines(to.show$homopolymer.length, exp(p$fit -  2* p$se.fit), col='red', lty=2)
lines(to.show$homopolymer.length, exp(p$fit), col='#FF00000d', lwd=2)
for(q in seq(0.025, 0.475,0.025))
  polygon(c(to.show$homopolymer.length, rev(to.show$homopolymer.length)), c(exp(qnorm(q, p$fit, p$se.fit)), exp(rev(qnorm(1-q, p$fit, p$se.fit)))), border=NA, col='#FF00000d')
```

Double mutants, **pole-4; pms-2**, generation 10
```{r, eval=F}
fit <- gam(cbind(pole4pms2, total_frequency-pole4pms2) ~ s(homopolymer.length, 6), data=to.show, family=binomial)
plot(to.show$homopolymer.length, to.show$pole4pms2/to.show$total_frequency, yaxt="n", xlim=c(4,18), ylim=c(0,0.015), ylab=NA, pch=16, col='grey', main='pole-4; pms-2',font.main=3, xlab="Homopolymer length")
segments(to.show$homopolymer.length, qbeta(0.025, .5+to.show$pole4pms2, .5+to.show$total_frequency- to.show$pole4pms2),
         to.show$homopolymer.length, qbeta(0.975, .5+to.show$pole4pms2, .5+to.show$total_frequency- to.show$pole4pms2), col='grey')
p <-  predict(fit, type='link', se.fit=TRUE)
lines(to.show$homopolymer.length, exp(p$fit), col='red', lwd=2)
lines(to.show$homopolymer.length, exp( p$fit +  2*p$se.fit), col='red', lty=2)
lines(to.show$homopolymer.length, exp(p$fit -  2* p$se.fit), col='red', lty=2)
lines(to.show$homopolymer.length, exp(p$fit), col='#FF00000d', lwd=2)
for(q in seq(0.025, 0.475,0.025))
  polygon(c(to.show$homopolymer.length, rev(to.show$homopolymer.length)), c(exp(qnorm(q, p$fit, p$se.fit)), exp(rev(qnorm(1-q, p$fit, p$se.fit)))), border=NA, col='#FF00000d')
```
